---
title: "My R Markdown notebook"
output: html_notebook
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. -->

This document will introduce you to R Markdown notebooks, and give you some hands-on experience adding to the document.

# Getting started

You can create a new R Markdown document in RStudio by clicking the menu item:

* File / New File >  R Notebook

_or_

* File / New File >  R Markdown...


For this course, we will be using the Notebook option...for your project, you may wish to use another format.

# The first chunk

It's good practice to load all of the packages you'll require at the top of your script. When a package is loaded, all of the functions and data in the package are available.

Loading an installed package is done with the `library()` function. The R chunk below will load the {tidyverse} package (which in turn will load other packages, such as the data manipulation package {dplyr} and the plotting package {ggplot2}). 

To run a chunk, press the green arrow at the far right of the chunk, or put your cursor in the chunk and press `Ctrl + Shift + Enter` (on macOS, `Cmd + Shift + Enter`) 

## Your Turn 1.1

Add the package {beepr} to the setup chunk, and run the chunk.



```{r setup}
library(tidyverse)

# solution
library(beepr)

```


## Your Turn 1.2

Add a new chunk to the R Markdown document.

You can do this with the "Insert" toolbar button, or the keyboard shortcut 

* `Ctrl + Alt + I` 

* `Cmd + Option + I` on macOS


You could also type the three back ticks, the curly braces, the letter R, and three more backticks to end the chunk--but that's a very WET approach!




## Your Turn 1.3

In that chunk, put the function `beep()` and run it...

Any Zelda fans? Try `beep("treasure)`!  You can find out the full range by typing ?beep into the console...

```{r}
# solution
beep()
```

Enough of that silliness...onward!


# 2. Chunk contents

## Your Turn 2.1

R code chunks can contain anything you want to do in R.

Here's a chunk that contains code to create a chart from everyone's favourite data set, `mpg`

Run the chunk and see what happens...


```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```


Code chunks can create an object, which is stored in the Global Environment...it will be listed under the "Environment" tab in the top right of your RStudio screen.

That object can then be referenced in a future chunk. This can be very useful!

## Your Turn 2.2

Use the `seq()` function to create an object "my_numbers" that contains the integers from 1 to 20.

If you're not sure about the syntax of the function, don't hesitate to use the ?seq command in the console...

```{r one_to_seven}
# solution
my_list <- seq(1:7)

```

Now, use the next chunk to calculate the sum of the numbers in `my_list`

```{r}
#solution
sum(my_list)
```


# 3. Inline text


### Your Turn 3.1
















# 9. Rendering the HTML file

At this point, we want to knit it all together.


## Your Turn 9.9 - rendering the HTML file

Here again is that chunk that creates a chart. When we are rendering the final document, we might not want to have the code displayed (it might be a fancy report for your boss's boss) -- we just want the chart.

The solution is to use one of many 


```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```

The solution is to put the option "echo = FALSE" after the "r" in the opening bit of the code.




## REFERENCE MATERIAL

You can find more code chunk options described here:
https://bookdown.org/yihui/rmarkdown/r-code.html

There's also a short listing of them in the R Markdown cheatsheet and reference guide, available online and through the "Help / Cheatsheets >" menu item in RStudio




