<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  

-->

# (PART) Week 3 {-} 

<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License
-->



# Introduction to data visualization {#data_viz}


```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)

library(tidyverse)
library(gapminder)
```




## Anscombe's Quartet


"Anscombe's Quartet" comprises four data sets that each have eleven rows, each with two variables (_x_ and _y_). The quartet was constructed by Francis Anscombe, and published in a 1973 paper titled ["Graphs in Statistical Analysis" in the journal _American Statistician_](https://www.sjsu.edu/faculty/gerstman/StatPrimer/anscombe1973.pdf)[@Anscombe_1973].

The quartet is included in base R, but in an untidy format:

```{r}
anscombe
```


Read the quartet data in a tidy form^[Source: David Robinson, ["Anscombe’s Quartet with Tidy Tools"](https://rpubs.com/dgrtwo/tidy-anscombe)]



```{r, echo = FALSE}
## old-school `gather()` and `spread()`
#anscombe_tidy <- anscombe %>%
#  mutate(observation = seq_len(n())) %>%
#  gather(key, value, -observation) %>%
#  separate(key, c("variable", "set"), 1, convert = TRUE) %>%
#  mutate(set = c("I", "II", "III", "IV")[set]) %>%
#  spread(variable, value) %>%
#  arrange(set)
#anscombe_tidy
#write_csv(anscombe_tidy, "anscombe_tidy.csv")
#
```


```{r anscombe_tidy}
anscombe_tidy <- read_csv("anscombe_tidy.csv")

anscombe_tidy
```



### 1. Summary statistics

Each of the four sets in Anscombe's Quartet has the same summary statistics. Let's calculate the mean of _x_ for each of the four sets:


```{r}
# example
anscombe_tidy %>%
  group_by(set) %>%
  summarize(mean.x = mean(x))
```

### Your turn 1.1

Using the following functions, calculate the summary statistics of _x_ and _y_, and the correlation coefficient between _x_ and _y_, for all four of the sets in the quartet:

 | statistic | function |
 | ----       | ----      |
 | mean      | mean()   |
 | standard deviation | sd() |
 | correlation coefficient | cor() |
 
 

```{r}
# solution
anscombe_tidy %>% 
  group_by(set) %>% 
  summarize(mean(x), 
            sd(x), 
            mean(y), 
            sd(y), 
            cor(x, y))
```


## 2. Visualizing the quartet

Using the R visualization package {ggplot2}.

The template of a `ggplot()` function call looks like this:

`ggplot(data = <DATA>) + ` <BR>
`  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`  


With this example, the dataframe we're using is `anscombe_tidy` and we are plotting `x` and `y` as points.

```{r}
# example
ggplot(anscombe_tidy) + 
  geom_point(aes(x = x, y = y)) 
```


### Your turn 2.1



With `colour` to differentiate each set

```{r}
# example
ggplot(anscombe_tidy) + 
  geom_point(aes(x = x, y = y, colour = set)) 
```


## Your turn 1.3

Filter by set

```{r}
# solution
anscombe_tidy %>%
  filter(set == "I") %>%
ggplot() + 
  geom_point(aes(x, y), size = 3, colour = "purple")
```




## Your turn 1.4

Facet plot

```{r}
ggplot(anscombe_tidy) + 
  geom_point(aes(x, y)) + 
  facet_wrap(~set)
```



## Your turn 1.5

Add a trend line using the `geom_smooth` function.

* the `method = lm` indicates a "linear model", i.e. a standard regression line. {ggplot2} provides access to other smoothing algorithms.

* the `se = FALSE` turns off the "standard error" (a measure of uncertainty in the data)

```{r}
ggplot(anscombe_tidy) + 
  geom_point(aes(x, y)) + 
  geom_smooth(aes(x, y), method = lm) +
  facet_wrap(~set)
```


But that duplicates the `aes(x, y)` text ... so we can move that into the `ggplot()` function. That way, the aesthetics apply to each of the `geom_` calls.


```{r}
p <- ggplot(anscombe_tidy, aes(x, y)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  facet_wrap(~set)
p
```

Try #E5E5E3 as background colour:
https://dataviztoday.com/blog/37

```{r}
p
p + theme(panel.background = element_rect(fill = "#E5E5E3"))
```



-30-
