# (PART) Week 5 {-} 



<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  

Some content is drawn from 

A. _R for Data Science_ by Hadley Wickham and Garrett Grolemund
-- (https://r4ds.had.co.nz/) -- which is licensed under the Creative Commons
Attribution-NonCommercial-NoDerivs 3.0 (CC BY-NC-ND 3.0 US) 


-->


# Dates and times {#date-time}


```{r echo = FALSE}

library(tidyverse)
library(gapminder)

library(lubridate)  # part of the tidyverse, but not one of the core packages that loads with {tidyverse}
```


## Objectives {#date-time-objective}

* Understand the basics of date and time formats

* Understand basic functions in {lubridate} for working with dates

* Demonstrate ability to use those functions



## Reading and reference {#date-time-reading}

["Dates and times" in _R for Data Science_](https://r4ds.had.co.nz/dates-and-times.html) by Hadley Wickham and Garrett Grolemund

[Dates and Times Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf) (PDF file)

[{lubridate}](https://lubridate.tidyverse.org/index.html) package reference

["Strings and Dates" in _R Cookbook_ (2nd ed.)](https://rc2e.com/stringsanddates)

[Working with dates and time in R using the lubridate package](https://data.library.virginia.edu/working-with-dates-and-time-in-r-using-the-lubridate-package/), University of Virginia Library Research Data Services + Sciences


**going beyond**

[{clock}: a brand-new package for working with date-times](https://clock.r-lib.org/)

* [Comprehensive Date-Time Handling for R](https://www.tidyverse.org/blog/2021/03/clock-0-1-0/).  "lubridate will never go away, and is not being deprecated or superseded. As of now, we consider clock to be an alternative to lubridate. "


## Introduction: Dates and Times {#date-time-intro}

Units of time:

Financial calendars

* [4-4-5 calendar](https://en.wikipedia.org/wiki/4%E2%80%934%E2%80%935_calendar): each quarter is divided into three "months", of 4 or 5 weeks length



## {lubridate} functions

See https://lubridate.tidyverse.org/reference/index.html 

### parsing functions 

| function | purpose |
| ---- | ---- |
| ymd()  | returns yyyy-mm-dd |
| dmy() | returns yyyy-mm-dd from day-month-year |
| _etc_ |  |
| ymd_hms() | returns date and time |


### getting, setting, rounding

| function | purpose |
| ---- | ---- |
| today() |  |
| now() |  |
| year(), month(), day() | get components of date|
| _etc_ |  |
| floor_date() | rounds _down_ to nearest boundary of specified time unit |
| ceiling_date() | rounds _up_ to nearest boundary of specified time unit |


## Parsing {#date-time-parsing}

{lubridate} will happily work with numbers and strings, including strings with separators

```{r}

ymd(20200328)

ymd("2020-03-28")

```



Parse "10-11-12" using `dmy()` ... what do you get back?

```{r}
# solution
dmy("10-11-12")

```


What about `mdy()` and `ymd()`?




## Date {#date-time-date}

The three functions `year()`, `month()`, and `day()` allow you to pull that component from a ymd value.

Here are three dates:

```{r}
date1 <- ymd("2019-01-02")
date2 <- ymd("2020-02-29")
date3 <- ymd("2020-11-21")

```

Here's how we would get the year from `date1`:

```{r}

year(date1)

```



Get the month from `date2`

```{r}
# solution

month(date2)

```

Get the day from `date3`

```{r}

day(date3)

```


What happens when you add the options `label` and `abbr` to the function you used to get the day?

* experiment with setting the options to TRUE or FALSE

```{r}
# solution
month(date2, label = TRUE, abbr = FALSE)

```



What does `wday()` do? 

* again, experiment with `label` and `abbr` 

```{r}
# solution

wday(date3)

my_wday <- wday(date3, label = TRUE, abbr = FALSE)
my_wday

class(my_wday)
class(month(date2, label = TRUE, abbr = FALSE))


```

Bonus: what class of variable is `wday` returning?


Another useful function is `week()`, which returns the week number. More precisely, 

> `week()` returns the number of complete seven day periods that have occurred between the date and January 1st, plus one.

We often say that "there are 52 weeks in a year", but 365 / 7 = 52.1. This means that the last few days of the year are in the 53rd week.

```{r}
library(lubridate)

# January 2nd, 2019
week(ymd("2019-01-02"))

# December 30, 2020
week(ymd("2020-12-30"))

```



## Time span {#date-time-span}

There are three classes that we can use:

> * durations, which represent an exact number of seconds.
> * periods, which represent human units like weeks and months.
> * intervals, which represent a starting and ending point.

(from _R for Data Science_)



The ideas of duration and interval are useful in some contexts, where you want to measure a period of time in seconds. For example, you might be measuring how long it takes an individual on a treadmill to get their heart rate up to 120 beats per minute.

For this exercise, we will keep using our three dates and measure the period between them.

We can **subtract** two dates:

```{r}

date2 - date1

```


Here's three different ways of **adding** a year to a date:

```{r}
start_date <- ymd("2019-01-01")

start_date + 365

start_date + dyears(1)

start_date + years(1)
```


2020 was a leap year... what happens if you use the same three equations that add a year but change the date to "2020-01-01"?

```{r}
# solution
start_date <- ymd("2020-01-01")

start_date + 365

start_date + dyears(1)

start_date + years(1)


```




## Working with dates {#date-time-working}


Read in this extract of a file downloaded from Statistics Canada, which has monthly data showing the British Columbia labour force (both sexes, 15 years and over, unadjusted) from January 2000 to October 2019.

```{r}

lfs_BC_2000 <- read_csv("data/lfs_BC_2000.csv")
lfs_BC_2000
```

The date is shown in the variable `REF_DATE` (short for "reference date"); since it's monthly data, the character representation contains the year and the month, but not the day.

What happens when you mutate `REF_DATE` into a new variable, using the `ymd()` function from {lubridate}?



```{r}
# solution
lfs_BC_2000 <-
  lfs_BC_2000 %>% 
  mutate(ref_date_2 = ymd(REF_DATE))

lfs_BC_2000
```


Oh no! 94 of the 238 failed to parse—the program was unable to convert them due to ambiguity in the numbers. These show up as "NA" values, such as "2000-01". {lubridate} is trying to find a year, a month, and a day in this string, and the "00" value doesn't make sense in this context.

And some of the months that did parse ended up with completely wrong interpretations. The "20" in "2001-01" is interpreted as the year, and the "01"s become the month and day, so it shows up as "2020-01-01".

Back to the drawing board...

Since the variable `REF_DATE` is a character string, a solution is to force them to all be the first day of the month. I'll do this by adding "-01" to the end of the string, using the `glue()` function (which is in the {glue} package.

* https://glue.tidyverse.org/

**NOTE:** 

* As with pretty much everything else, there is more than one way to solve this problem!

**NOTE:** 

* You can use a function that you have installed but not loaded through the `library()` function, by putting the name of the package followed by the function you want to use. In this case, we are using the `glue()` function from the {glue} package, so our code looks like `glue::glue()`. The variable from our data frame gets put in braces, and the string returned is all within quotation marks.


```{r}

lfs_BC_2000 <-lfs_BC_2000 %>% 
  mutate(ref_date_2 = glue::glue("{REF_DATE}-01"))

lfs_BC_2000
```



Now convert `ref_date_2` using `ymd` ...

```{r}
# solution

lfs_BC_2000 <- lfs_BC_2000 %>% 
  mutate(ref_date_2 = ymd(ref_date_2))

lfs_BC_2000
```

With the reference date turned into a date format, we can create a line plot, and {ggplot2} will make some decisions for us on the format of the labels on the x-axis.

```{r}

ggplot(lfs_BC_2000, aes(x = ref_date_2, y = VALUE)) +
  geom_line()

```



And now we will create a summary table...the annual average of the number of people in the B.C. labour force.

In this code chunk, the variable "ref_date_2" is inside the `year()` function. This means that the year becomes the grouping variable.

```{r}

lfs_BC_2000 %>%
  group_by(year(ref_date_2)) %>% 
  summarise(annual_average = mean(VALUE))

```


Looks like the B.C. labour force is growing!



Is there any seaonality in the size of the labour force? To calculate the **monthly** average over the 20 years in our data, we can use the `month()` function to use the 12 months of the year as our grouping variable.


```{r}

# solution 1
lfs_BC_2000 %>%
  group_by(month(ref_date_2)) %>% 
  summarise(monthly_average = mean(VALUE))
```

In this second solution, the `mutate()` function is used to create a stand-alone variable "ref_month", which becomes the grouping variable.

The `month()` function also includes the argument `label = TRUE`, which adds the abbreviations for the month names.

```{r}

# solution 2
lfs_BC_2000 %>%
  mutate(ref_month = month(ref_date_2, label = TRUE)) %>% 
  group_by(ref_month) %>% 
  summarise(monthly_average = mean(VALUE))

```



Would a simple data visualization (a plot) help you see if there is a pattern?

Of course the answer is "Yes!" Visualization is an important tool in exploratory data analysis. But you knew that already.

(Note that if you haven't already, you will have to create a separate variable with the month value. As of this writing, {ggplot2} doesn't allow you to run a function inside the `aes()`.)

```{r}

lfs_BC_2000 %>%
  mutate(ref_month = month(ref_date_2)) %>% 
  group_by(ref_month) %>% 
  summarise(monthly_average = mean(VALUE)) %>% 
  ggplot(aes(x = ref_month, y = monthly_average)) +
  geom_line()


```


Our plot is a bit off, though, with the x-axis labels showing decimal place months. This is because the "ref_month" variable is a numeric value. 
An effective solution is specify the type of scale in the x-axis, using `scale_x_discrete()`.


```{r}

lfs_BC_2000 %>%
  mutate(ref_month = month(ref_date_2)) %>% 
  group_by(ref_month) %>% 
  summarise(monthly_average = mean(VALUE)) %>% 
  ggplot(aes(x = ref_month, y = monthly_average)) +
  geom_line() +
  # add x-axis scale
  scale_x_discrete(limits = month.abb)

```


Important note: leaving the `labels = TRUE` will not work with this plot! That's because it becomes a factor, which can't be plotted as a line. One solution would be to plot the values using `geom_col()`.


## Date-time {#date-time-date-time}

What is the difference between `today()` and `now()`?

```{r}
# solution

today()

now()

```



You can specify time zones by adding a time zone option to your call. For example, our time zone is often represented as "America/Los_Angeles" or "America/Vancouver".

```{r}

Sys.timezone()

```




What time is it right now? Specify the time zone in the `now()` function with the argument `tzone =` or `tz =`

* here in Victoria -- or maybe Vancouver

* in Seoul, South Korea

* in Lima, Peru

* and Universal Time Coordinated (the time zone is represented as "UTC")


```{r}

now(tzone = "America/Vancouver")

now(tzone = "America/Lima")

now(tz = "Asia/Seoul")

now(tz = "UTC")

```


You can use the function `OlsonNames()` to print a list of all 425 time zones around the world!

```{r}
# solution -- beware the length!
head(OlsonNames())
```


(Note: wrapping a function call with an assignment both runs the assignment _and_ prints the object!)

```{r }
(x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York"))

(x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen"))

(x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland"))

```


Subtract date-time from each other

```{r}
#
x2 - x3

```

Different ways to represent March 28, 2020:

March 28, 2020
2020-03-28
28 Mar 2020
28-3-20
3/28/2020

Also, keep in mind that a written-out-month is spelled differently in different languages! For example, April in French is "Avril".



## Lead and lag {#date-time-lead-lag}

```{r}

df_dates <- tribble(~event, ~event_date,
        "event_1", "2015-01-15",
        "event_2", "2020-02-29",
        "event_3", "2020-04-15"
        )

df_dates <- df_dates %>% 
  mutate(event_date = ymd(event_date))

df_dates

df_summary <- df_dates %>% 
  # create a new variable with the lag from the previous record
  mutate(lag_date = lag(event_date)) %>% 
  # calculate the different between the two dates
  mutate(days_between = event_date - lag_date)

df_summary

# calculate the number of years (instead of days)
# - note the need to convert "days_between" from Duration to integer
#   before dividing by 365.25
df_summary %>% 
  mutate(years_between = as.integer(days_between)/365.25)

df_summary %>% 
  summarise(avg_days_between_events = mean(days_between, na.rm = TRUE))

```



-30-
