<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  

-->

```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)
```

# Communication: table formatting {#communication-table}



## Reading & reference  {#communication-table-reading}

Garrett Grolemund and Hadley Wickham, [R for Data Science](https://r4ds.had.co.nz/)

* ["Communication" section](https://r4ds.had.co.nz/communicate-intro.html), Chapters 26 "Introduction" to 29 "R Markdown formats"




**R Markdown**

Yihui Xie, Christophe Dervieux, Emily Riederer, ["The kableExtra package", _R Markdown Cookbook_](https://bookdown.org/yihui/rmarkdown-cookbook/kableextra.html)



Packages:

**kableExtra**

* Hao Zhu, [Create Awesome HTML Table with knitr::kable and kableExtra](https://haozhu233.github.io/kableExtra/awesome_table_in_html.html)

* Hao Zhu, [Using kableExtra in Bookdown](https://haozhu233.github.io/kableExtra/bookdown/index.html)

* Yihui Xie, Christophe Dervieux, Emily Riederer, ["The kableExtra package", _R Markdown Cookbook_](https://bookdown.org/yihui/rmarkdown-cookbook/kableextra.html)

* [KableExtra Tutorial](https://rstudio-pubs-static.s3.amazonaws.com/444395_76727eaf9c774fa8bf932bed16500a00.html)




**flextable**

* {flextable} - [flextable R package reference](https://davidgohel.github.io/flextable/index.html)

* David Gohel, [flextable book](https://ardata-fr.github.io/flextable-book/)


## Introduction {#communication-table-intro}

The look of the default table in R markdown is not all that attractive. Here's a summary table of the Oakland Athletics' seasons from 1999-2008.

```{r moneyball_oakland}

oakland_99 <- mlb_pay_wl %>% 
  filter(tm == "OAK" & year_num < 2009)

oakland_99
```

There are a few R packages that allow us to format a table like this, ready for publication. Here we will take a quick look at two of them. Both have many, many more options for formatting; see the reference materials at the top of this chapter for those details.


## The kable package {#communication-table-kable}


First, {kableExtra} is an extension to the {kable} tool that is part of "knitting" our R Markdown files.

```{r}
library(kableExtra)

# print table with {kableExtra} formatting  
oakland_99 %>%
  # create table, add title
  kable(caption = "Oakland Athletics, 1999-2008") %>%
  kable_styling(bootstrap_options = "striped", font_size = 10) %>%
  # make variable names bold and large
  row_spec(0, bold = T, font_size = 14) %>%
  # make 2002 season (row 4) bold
  row_spec(4, bold = T) 
```  



## The flextable package {#communication-table-flextable}


Another table formating package is {flextable}. 


```{r}
library(flextable)

oakland_99 %>% 
  flextable() %>% 
  add_footer_row(values = "Source: baseball-reference.com", colwidths = 8)

```


-30-
