<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  
-->

# (PART) Week 6 {-} 


# Where to from here? {#where-to}


```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)

```


You've just started your journey into the world of data analysis in a coding environment, and now know some methods for working in the programming language R (and even a bit of Python).

So...where to from here? 

With the foundation you now have, you can go in a variety of directions. And remember, they are not mutually exclusive! And this is not an exhaustive list.



## More _R for Data Science_ and beyond {#where-to-more-r4ds}

We didn't cover everything in _R for Data Science_. Here's a short summary of those topics, and some next-steps to go beyond what's in the book.

### R for Data Science book club

There is an online book club that works its way through _R for Data Science_. If you're interested in a combination of a refresher of BIDA302 and a deeper dive into some of the topics (see below), this self-directed and supported learning might be for you.

Information about the book club can be found at the GitHub repo [github.com/r4ds/bookclub-R_for_Data_Science](https://github.com/r4ds/bookclub-R_for_Data_Science).


### Functional programming

> R, at its heart, is a functional programming (FP) language. This means that it provides many tools for the creation and manipulation of functions.
–Hadley Wickham, _Advanced R_ [@Wickham_advancedR]

R, and specifically the package {purrr}, supports a concept called "functional programming". This means that things like for-loops can be turned into functions; the {purrr} package has functions that do just that.

See ["Iteration" in _R for Data Science_, starting at 21.4 "For loops vs. functionals"](https://r4ds.had.co.nz/iteration.html)

When we learned the concept of a for-loop earlier, we combined three Excel files. Here's how we would do that using the `map_()` functions within the {purrr} package, with the added twists of: 

* getting all the data from all the sheets as well as all the files, 

* creating variables with the file name and the sheet name.


```{r}
all_files <- dir("data_monthly/", pattern = "\\.xls*", full.names = TRUE)

result <- map_df(all_files, function(x) {
             #Get all the sheet names
             all_sheets <- readxl::excel_sheets(x)  
             #read the excel file with one sheet at a time
             map_df(all_sheets, ~readxl::read_excel(x, sheet = .x) %>% 
                       #add columns for filename and sheetname
                       dplyr::mutate(filename = basename(x), sheetname = .x))
})

```

To go even deeper into programming with R, I would suggest the book _Advanced R_ by Hadley Wickham [@Wickham_advancedR_2] 

* [online version of _Advanced R_](https://adv-r.hadley.nz/)


### Modeling

We covered two of the four chapters on statistical modeling that are included in _R for Data Science_. There's a lot more!

See ["Part IV Model" in _R for Data Science_, starting at 22 "Model: Introduction"](https://r4ds.had.co.nz/model-intro.html)


To go still further with modeling, you may wish to explore [tidymodels](https://www.tidymodels.org/):

> The tidymodels framework is a collection of packages for modeling and machine learning using tidyverse principles.

Some other texts dealing with modeling methodologies using R include:

* Scott Cunningham, _Causal Inference: The Mixtape_ [@Cunningham_mixtape]

  - [online version of _Causal Inference_](https://mixtape.scunning.com/)

* Richard McElreath, _Statistical Rethinking: A Bayesian Course with Examples in R and Stan_ [@McElreath_2016]

* Gareth James, Daniela Witten, Trevor Hastie, and Robert Tibshirani, _An Introduction to Statistical Learning_ [@James_Witten_Hastie_Tibshirani_2014]

  - [pdf version of _An Introduction to Statistical Learning_](https://static1.squarespace.com/static/5ff2adbe3fe4fe33db902812/t/6062a083acbfe82c7195b27d/1617076404560/ISLR%2BSeventh%2BPrinting.pdf)


-30-