# (PART) REFERENCE {-} 

<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  
-->


# Utilities {#utilities}


```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)

library(tidyverse)
library(janitor)
```

One of the many great things about the R ecosystem is that data scientists around the world are tackling the same types of things that you and I might encounter, creating solutions to those problems, and then sharing those solutions as packages that add new functionality to the ecosystem.

This chapter introduces some of the packages that I have found useful.





## {datapasta}  ·

Reference page: https://milesmcbain.github.io/datapasta/

> datapasta is about reducing resistance associated with copying and pasting data to and from R. It is a response to the realisation that I often found myself using intermediate programs like Sublime to munge text into suitable formats. Addins and functions in datapasta support a wide variety of input and output situations, so it (probably) “just works”. Hopefully tools in this package will remove such intermediate steps and associated frustrations from our data slinging workflows.

We don't use {datapasta} in this course, but the chances are high that someday soon you'll need to copy-and-paste data.



## `{here}`  ·

Reference page: https://here.r-lib.org/

> The goal of the here package is to enable easy file referencing in project-oriented workflows. In contrast to using setwd(), which is fragile and dependent on the way you organize your files, here uses the top-level directory of a project to easily build paths to files.

The `here()` package is used as part of the Workflow chapter.



## `{janitor}`  ·

Reference page: https://garthtarr.github.io/meatR/janitor.html

> The janitor package is a R package that has simple functions for examining and cleaning dirty data. It was built with beginning and intermediate R users in mind and is optimised for user-friendliness.

> The main janitor functions:

>* perfectly format data frame column names;
>
>* isolate partially-duplicate records; and
>
>* provide quick tabulations (i.e., frequency tables and crosstabs).


| function | action |
| :-- | :-- |
| **cleaning** |  |
| `clean_names()` | handles problematic variable names |
| `remove_empty()` | removes any columns that are entirely empty and entire rows that are entirely empty |
| `excel_numeric_to_date()` | converts Excel date serial numbers to class `Date` |
| **data summaries** |  |
| `tabyl()` |  takes a vector and returns a frequency table |
| `adorn_()` | a family of functions for additional summary values |


```{r, evaluate=FALSE}
library(janitor)
```

This is going to demonstrate {janitor} `clean_names()` function

First we read in the valid rows from the "deaths.xlsx" file (part of the [{readxl} package] https://readxl.tidyverse.org/).

```{r}
deaths <- read_excel("deaths.xlsx", skip = 4, n_max = 10)

deaths
```

If we wanted to filter the cases where the individual had children, we have to surround the variable name in backticks, since there is a space. We also have to remember to capitalize the "H" (although this is less of an issue, as RStudio's autofill is not sensitive to case).

```{r}
deaths %>% 
  filter(`Has kids` == TRUE)
```

Using the `clean_names()` function from {janitor} gives us a solution, in that it makes the variable names consistently lower-case, and with the spaces replaced by an underscore ("_"). 

```{r}
deaths <- deaths %>% 
  janitor::clean_names()

deaths
```

This then means that our filter statement does not need backticks around the variable name.

```{r}
deaths %>% 
  filter(has_kids == TRUE)
```

{janitor} has a range of other useful functions, including removing empty columns and rows, and creating summary cross-tab tables.



## But wait, there's more!

Here's a twitter thread on packages that people nominated as deserving more attention:

https://twitter.com/WeAreRLadies/status/1369984394836910084
