---
title: "Utilities"
author: "Martin Monkman"
output:
  html_document
  
---

```{r setup}
# bonus utility - {pacman}
# https://github.com/trinker/pacman
pacman::p_load(tidyverse, lubridate, gapminder)
#
# utilities
library(janitor)
library(datapasta)
library(glue)
```




## Utilities in R


## 1. {glue} - 

The {glue} package provides some useful tools for sticking together character strings and data.

* [{glue}](https://glue.tidyverse.org/)

* You can use a function that you have installed but not loaded through the `library()` function, by putting the name of the package followed by the function you want to use. In this case, we are using the `glue()` function from the {glue} package, so our code looks like `glue::glue()`. The variable from our data frame gets put in braces, and the string returned is all within quotation marks.


### `glue()`  

Here's a simple example:

```{r}

glue('BIDA', '302')

```

We can also paste together variables, by putting the name inside curly braces `{}`:

```{r}

course <- "BIDA"
course_number <- 301

glue({course}, {course_number})

```

We can specify a character to separate the elements we are glueing, with `.sep = "<string>"`. In this example it's a space: 

```{r}

glue('_The Mandolorian_',
  'is a tv series on the Disney+ streaming service',
  'that featured Baby Yoda.', .sep = " ")

```

### 1.1 Your turn

Glue together separate year (as a variable), month, and day values, separated by a hyphen "-", to create a single date string.

```{r}
# assign year to variable
this_year <- 2020

# solution

```






### `glue_data()`  


We can glue data in a {dplyr} pipe with the `glue_data()` function. 

First, let's filter the {gapminder} data so we have a smaller piece to work with:

```{r}

gm_3 <- gapminder %>%
  filter(year == 2007,
         country %in% c("Canada", "Japan", "New Zealand"))

gm_3

```

Now we can create multiple strings:

```{r}

gm_3 %>%
  glue_data("{country} has a life expectancy of {lifeExp}.")

```


We can also apply functions within the curly braces:

```{r}

gm_3 %>%
  glue_data("{country} has a life expectancy of {round(lifeExp, 1)}.")

```


### 1.2 Your turn

Introducing the `starwars` character data table:

```{r}

head(starwars)

```


Create a new variable in the data frame by glueing the variables `species` and `gender`.

```{r}
# solution


```



## 2. {janitor} -

Some simple functions for examining and cleaning data.

* [{janitor}](http://sfirke.github.io/janitor/)

  - [#twofunctionsmostdays: janitor](https://github.com/sharlagelfand/twofunctionsmostdays/tree/master/2020/03/16)

Let's look at a very messy Excel file. (This is all-too typical of the sorts of things you see in real life.) Some problems with this:

* Dreadful column names

* Rows and columns containing Excel formatting but no data

* Dates stored as numbers

* Values spread inconsistently over the “Certification” columns


```{r}

roster_raw <- readxl::read_excel("dirty_data.xlsx") # available at http://github.com/sfirke/janitor
glimpse(roster_raw)

roster_raw

```


```{r}
roster <- roster_raw %>%
  clean_names() %>%
  remove_empty(c("rows", "cols")) 

roster
```

We can also extend the pipe with other cleaning functions. Note that `excel_numeric_to_date()` is another {janitor} function, while `coalesce()` is from {dplyr} (and is similar to the SQL function of the same name).

```{r}
roster <- roster_raw %>%
  clean_names() %>%
  remove_empty(c("rows", "cols")) %>%
  mutate(hire_date = excel_numeric_to_date(hire_date),
         cert = coalesce(certification_9, certification_10)) %>% # from dplyr
  select(-certification_9, -certification_10) # drop unwanted columns

roster
```


* `coalesce()` finds the first non-missing value at each position

```{r}

x <- c(1, 2, NA, 4)
y <- c(NA, NA, 93, 94)

coalesce(x, y)

```

### 2.1.1 Your turn

What happens if you put `y` first in the `coalesce()` function?


```{r}
# solution


```


### duplicate records

We can find the cases where there are duplicates in the data records. In this case, we are looking for duplicate names:

```{r}
roster %>% get_dupes(first_name, last_name)
```


### 2.2 Your turn


Find the cases where the subject is duplicated.

```{r}
# solution



```


### simple tables

The {janitor} package also offers some simple tabulations with the `tabyl()` function.

One variable returns counts and percentages:

```{r}
roster %>%
  tabyl(full_time)
```


If we use two variables, it takes the first one as the rows, and the second as the columns:

```{r}

roster %>%
  tabyl(full_time, subject)

  
```  


And for three variables, it makes separate tables based on the third one, then creates tables like the two-variable version:

```{r}

roster %>%
  tabyl(full_time, subject, employee_status)

  
```  

There's also an option to drop levels with zero cases:

```{r}

roster %>%
  tabyl(full_time, subject, employee_status, show_missing_levels = FALSE)

  
```  


You can also add totals and percentages with the `adorn_()` functions. This example creates a fancy version of what we had before.

* Note: I found the naming of the totals a bit counter-intuitive. `adorn_totals("row")` creates a new row with the total for that column, while `adorn_totals("col")` is a new column with the row total.
 
```{r}
roster %>%
  tabyl(employee_status, full_time) %>%
  adorn_totals("col") %>%  
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title("combined")
```


See the [{janitor} reference page](http://sfirke.github.io/janitor/) for additional details on the table functions.



## 3. {datapasta} -

* [{datapasta}](https://milesmcbain.github.io/datapasta/)

While importing data stored as a csv or Excel file is (sort of) straight-forward in R, you might find data in another format, perhaps on a website or in a table in a Word document--{datapasta} can help. 

### Tangent - Addins in RStudio

* [RStudio addins, or how to make your coding life easier](https://towardsdatascience.com/rstudio-addins-or-how-to-make-your-coding-life-easier-6b627c7b2240)


### Example: Victoria climate norms


* [Victoria (UVic)](https://climate.weather.gc.ca/climate_data/daily_data_e.html?StationID=6812)

Step 1: go to the webpage, select the table, and <Ctrl>-c to copy the contents to your computer's clipboard

Step 2: in RStudio, put your cursor in an R chunk

Step 3: Addins > DATAPASTA > Paste as tribble


```{r}

tibble::tribble(
        ~Jan,       ~Feb,       ~Mar,       ~Apr,       ~May,       ~Jun,       ~Jul,       ~Aug,       ~Sep,       ~Oct,       ~Nov,       ~Dec, ~Year, ~Code,
       "4.6",      "5.1",      "6.8",      "9.0",     "12.1",     "14.9",     "16.9",     "16.8",     "14.2",     "10.0",      "6.4",      "4.0",    10,   "A",
       "1.4",      "1.5",      "1.1",      "0.9",      "1.0",      "0.9",      "0.9",      "0.7",      "0.8",      "0.7",      "1.6",      "1.4",   0.5,   "A",
       "7.6",      "8.8",     "10.8",     "13.6",     "16.9",     "19.9",     "22.4",     "22.4",     "19.6",     "14.2",      "9.7",      "7.0",  14.4,   "A",
       "1.5",      "1.3",      "2.6",      "4.3",      "7.2",      "9.8",     "11.3",     "11.1",      "8.6",      "5.7",      "3.0",      "1.1",   5.6,   "A",
      "16.1",     "18.3",     "21.4",     "26.3",     "31.5",     "33.5",     "36.3",     "34.4",     "31.1",     "27.6",     "18.3",     "16.1",    NA,    NA,
  "2005/ 19", "1963/ 10", "1994/ 28", "1998/ 30", "1983/ 28", "2003/ 06", "2007/ 11", "1960/ 09", "1944/ 05", "1987/ 01", "1975/ 03", "1940/ 21",    NA,    NA,
     "-15.6",    "-15.0",    "-10.0",     "-3.9",     "-1.1",      "2.1",      "4.1",      "4.4",     "-1.1",     "-4.4",    "-13.3",    "-14.4",    NA,    NA,
  "1950/ 28", "1950/ 01", "1989/ 03", "1956/ 05", "1954/ 01", "1991/ 03", "1979/ 02", "1973/ 18", "1972/ 27", "1956/ 28", "1955/ 14", "1964/ 16",    NA,    NA
  )


```

Dang it, it misses the row names--but we have something to start with.


```{r}

YYJ_climate_paste <- 
tibble::tribble(
        ~Jan,       ~Feb,       ~Mar,       ~Apr,       ~May,       ~Jun,       ~Jul,       ~Aug,       ~Sep,       ~Oct,       ~Nov,       ~Dec, ~Year, ~Code,
       "4.6",      "5.1",      "6.8",      "9.0",     "12.1",     "14.9",     "16.9",     "16.8",     "14.2",     "10.0",      "6.4",      "4.0",    10,   "A",
       "1.4",      "1.5",      "1.1",      "0.9",      "1.0",      "0.9",      "0.9",      "0.7",      "0.8",      "0.7",      "1.6",      "1.4",   0.5,   "A",
       "7.6",      "8.8",     "10.8",     "13.6",     "16.9",     "19.9",     "22.4",     "22.4",     "19.6",     "14.2",      "9.7",      "7.0",  14.4,   "A",
       "1.5",      "1.3",      "2.6",      "4.3",      "7.2",      "9.8",     "11.3",     "11.1",      "8.6",      "5.7",      "3.0",      "1.1",   5.6,   "A",
      "16.1",     "18.3",     "21.4",     "26.3",     "31.5",     "33.5",     "36.3",     "34.4",     "31.1",     "27.6",     "18.3",     "16.1",    NA,    NA,
  "2005/ 19", "1963/ 10", "1994/ 28", "1998/ 30", "1983/ 28", "2003/ 06", "2007/ 11", "1960/ 09", "1944/ 05", "1987/ 01", "1975/ 03", "1940/ 21",    NA,    NA,
     "-15.6",    "-15.0",    "-10.0",     "-3.9",     "-1.1",      "2.1",      "4.1",      "4.4",     "-1.1",     "-4.4",    "-13.3",    "-14.4",    NA,    NA,
  "1950/ 28", "1950/ 01", "1989/ 03", "1956/ 05", "1954/ 01", "1991/ 03", "1979/ 02", "1973/ 18", "1972/ 27", "1956/ 28", "1955/ 14", "1964/ 16",    NA,    NA
  )


```


So let's make a one-column tibble with the names of the variables. This is just a copy-paste from the web page, one item at a time.

```{r}

varnames <- tribble(~var_name,
                    "Daily Average (°C)",
                    "Standard Deviation",
                    "Daily Maximum (°C)",
                    "Daily Minimum (°C)",
                    "Extreme Maximum (°C)",
                    "Extreme Maximum - Date (yyyy/dd)",
                    "Extreme Minimum (°C)",
                    "Extreme Minimum - Date (yyyy/dd)"
)

```

Then we use the base R function `cbind()` (for "column bind"; there's a parallel `rbind()` function for adding rows to a table) to join the two objects.

```{r}

YYJ_climate_raw <- cbind(varnames, YYJ_climate_paste)

YYJ_climate_raw
```

Time to tidy the data--the observation is the month, not the variable.

Let's drop the two columns at the far right (the annual average and the code), then pivot it to longer and back wider but with the month as the observation.

```{r}

YYJ_climate <- YYJ_climate_raw %>% 
  select(-Year, -Code) %>% 
  pivot_longer(-var_name, names_to = "month", values_to = "value") %>% 
  pivot_wider(names_from = var_name, values_from = value) %>% 
  mutate(`Daily Average (°C)` = as.numeric(`Daily Average (°C)`)) %>% 
  mutate(month = as_factor(month))

YYJ_climate

```


And a plot of daily average temperature!

```{r}

ggplot(YYJ_climate, aes(x = month, y = `Daily Average (°C)`)) +
  geom_point()

```


-30-

