---
title: "Dates and times"
subtitle: "hands-on examples, with answers"
output: html_document
---

<!-- 
This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License (https://creativecommons.org/licenses/by/4.0/). 

Some content is drawn from 

A. _R for Data Science_ by Hadley Wickham and Garrett Grolemund
-- (https://r4ds.had.co.nz/) -- which is licensed under the Creative Commons
Attribution-NonCommercial-NoDerivs 3.0 (CC BY-NC-ND 3.0 US) 


-->

```{r setup}

library(tidyverse)
#
library(lubridate)  # part of the tidyverse, but not one of the core packages that loads with {tidyverse}

```



## Objectives

* Understand the basics of date and time formats

* Understand basic functions in {lubridate} for working with dates

* Demonstrate ability to use those functions



## {lubridate} functions

See https://lubridate.tidyverse.org/reference/index.html 

### parsing functions 

| function | purpose |
| ---- | ---- |
| ymd()  | returns yyyy-mm-dd |
| dmy() | returns yyyy-mm-dd from day-month-year |
| _etc_ |  |
| ymd_hms() | returns date and time |


### getting, setting, rounding

| function | purpose |
| ---- | ---- |
| today() |  |
| now() |  |
| year(), month(), day() | get components of date|
| _etc_ |  |
| floor_date() | rounds _down_ to nearest boundary of specified time unit |
| ceiling_date() | rounds _up_ to nearest boundary of specified time unit |


## 1. Parsing

{lubridate} will happily work with numbers and strings, including strings with separators

```{r}

ymd(20200328)

ymd("2020-03-28")

```


## Your turn 1.1 

Parse "10-11-12" using `dmy()` ... what do you get back?

```{r}
# solution
dmy("10-11-12")

```




## 2. Date 

The three functions `year()`, `month()`, and `day()` allow you to pull that component from a ymd value.

Here are three dates:

```{r}
date1 <- ymd("2019-11-02")
date2 <- ymd("2019-12-04")
date3 <- ymd("2020-02-29")

```

Here's how we would get the year from `date1`:

```{r}

year(date1)

```


## Your turn 2.1

Get the month from `date2`

```{r}
# solution

month(date2)

```

Get the day from `date3`

```{r}

day(date3)

```


What happens when you add the options `label` and `abbr` to the function you used to get the day?

* experiment with setting the options to TRUE or FALSE

```{r}
# solution
month(date2, label = TRUE, abbr = FALSE)

```


## Your turn 2.2

What does `wday()` do? 

* again, experiment with `label` and `abbr` 

```{r}
# solution

wday(date3)

my_wday <- wday(date3, label = TRUE, abbr = FALSE)
my_wday

class(my_wday)
class(month(date2, label = TRUE, abbr = FALSE))


```

Bonus: what class of variable is `wday` returning?


## 3. Time span

There are three classes that we can use:

> * durations, which represent an exact number of seconds.
> * periods, which represent human units like weeks and months.
> * intervals, which represent a starting and ending point.

(from _R for Data Science_)



The ideas of duration and interval are useful in some contexts, where you want to measure a period of time in seconds. For example, you might be measuring how long it takes an individual on a treadmill to get their heart rate up to 120 beats per minute.

For this exercise, we will keep using our three dates and measure the period between them.

We can **subtract** two dates:

```{r}

date2 - date1

```


Here's three different ways of **adding** a year to a date:

```{r}
start_date <- ymd("2019-01-01")

start_date + 365

start_date + dyears(1)

start_date + years(1)
```

## Your turn 3.1

2020 is a leap year... what happens if you use the same three equations that add a year but change the date to "2020-01-01"?

```{r}
# solution
start_date <- ymd("2020-01-01")

start_date + 365

start_date + dyears(1)

start_date + years(1)


```



---

## 4. Working with dates

### Your turn 4.1


Read in this extract of a file downloaded from Statistics Canada, which has monthly data.

```{r}

lfs_BC_2000 <- read_csv("lfs_BC_2000.csv")
lfs_BC_2000
```

The date is shown in the variable `REF_DATE`; since it's monthly data, the character representation contains the year and the month, but not the day.

What happens when you mutate `REF_DATE` into a new variable, using the `ymd()` function?

### Your turn 4.2


```{r}
# solution
lfs_BC_2000 <-
  lfs_BC_2000 %>% 
  mutate(ref_date_2 = ymd(REF_DATE))

lfs_BC_2000
```


Oh no! 94 of the 238 failed to parse--the program was unable to convert them due to ambiguity in the numbers. Back to the drawing board...

Since the variable `REF_DATE` is a character string, I am going to force them to all be the first day of the month. I'll do this by adding "-01" to the end of the string, using the `glue()` function (which is in the {glue} package.

* https://glue.tidyverse.org/

**NOTE:** 

* As with pretty much everything else, there is more than one way to solve this problem!

**NOTE:** 

* You can use a function that you have installed but not loaded through the `library()` function, by putting the name of the package followed by the function you want to use. In this case, we are using the `glue()` function from the {glue} package, so our code looks like `glue::glue()`. The variable from our data frame gets put in braces, and the string returned is all within quotation marks.


```{r}

lfs_BC_2000 <-lfs_BC_2000 %>% 
  mutate(ref_date_2 = glue::glue("{REF_DATE}-01"))

lfs_BC_2000
```


### Your turn 4.3

Now convert `ref_date_2` using `ymd` ...

```{r}
# solution

lfs_BC_2000 <- lfs_BC_2000 %>% 
  mutate(ref_date_2 = ymd(ref_date_2))

lfs_BC_2000
```


And now a summary table...the annual average of the number of people in the B.C. labour force:

```{r}

lfs_BC_2000 %>%
  group_by(year(ref_date_2)) %>% 
  summarise(annual_average = mean(VALUE))

lfs_BC_2000 %>%
  filter(year(ref_date_2) >= 2015)

```


Looks like the B.C. labour force is growing!


### Your turn 4.4

Is there any seaonality in the size of the labour force? Calculate the **monthly** average over the 20 years in our data:

```{r}

# solution 1
lfs_BC_2000 %>%
  group_by(month(ref_date_2)) %>% 
  summarise(monthly_average = mean(VALUE))


# solution 2
lfs_BC_2000 %>%
  mutate(month = month(ref_date_2)) %>% 
  group_by(month) %>% 
  summarise(monthly_average = mean(VALUE))

```


### Your turn 4.5

Would a simple data visualization (a plot) help you see if there is a pattern?

Of course the answer is "Yes!" Visualization is an important tool in exploratory data analysis. But you knew that already.

(Note that if you haven't already, you will have to create a separate variable with the month value.)

```{r}

lfs_BC_2000 %>%
  mutate(month = month(ref_date_2)) %>% 
  group_by(month) %>% 
  summarise(monthly_average = mean(VALUE)) %>% 
  ggplot(aes(x = month, y = monthly_average)) +
  geom_line()


```





---


## Date-time



## Your Turn 1.1

What is the difference between `today()` and `now()`?

```{r}
# solution

today()

now()

```


## Your Turn 1.2

You can specify time zones by adding a time zone option to your call. For example, our time zone is often represented as "America/Los_Angeles" or "America/Vancouver".

```{r}

Sys.timezone()

```




What time is it right now?

* here in Victoria -- or maybe Vancouver

* in Seoul, South Korea

* and Universal Time Coordinated (the time zone is represented as "UTC")


```{r}

now(tzone = "America/Vancouver")

now(tzone = "Asia/Seoul")

now(tzone = "UTC")

```


You can use the function `OlsonNames()` to print a list of all 425 time zones around the world!

```{r}
# solution -- beware the length!
head(OlsonNames())
```


(Note: wrapping a function call with an assignment both runs the assignment _and_ prints the object!)

```{r }
(x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York"))

(x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen"))

(x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland"))

```


Subtract date-time from each other

```{r}
#
x2 - x3

```

Different ways to represent March 28, 2020:

March 28, 2020
2020-03-28
28 Mar 2020
28-3-20
3/28/2020

Also, keep in mind that a written-out-month is spelled differently in different languages! For example, April in French is "Avril".


## REFERENCE

["Dates and times" in _R for Data Science_](https://r4ds.had.co.nz/dates-and-times.html) by Hadley Wickham and Garrett Grolemund

[Dates and Times Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/lubridate.pdf) (PDF file)

[{lubridate}](https://lubridate.tidyverse.org/index.html) package reference

[Working with dates and time in R using the lubridate package](https://data.library.virginia.edu/working-with-dates-and-time-in-r-using-the-lubridate-package/), University of Virginia Library Research Data Services + Sciences

