---
title: "BIDA302 - assignment #3"
subtitle: "solutions"
output: html_notebook
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
library(readxl)
#
library(gapminder)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


### 1. 
> (3 marks)
>
> a. Create a scatterplot using the {gapminder} data, showing all the countries in the world for 2007, the per capita GDP on the _x_ axis, and life expectancy on the _y_
>

 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

```

> b. Create a variant of the above plot, with each continent coloured differently

 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()

```


> c. And now using multiple facets
 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  facet_grid(~continent)

```


### 2. 
> (5 marks)
>


**Answer:**

> a. 


```{r}
# solution


```

> b. 

```{r}
# solution


```

