---
title: "BIDA302 - assignment #3"
subtitle: "solutions"
output: html_notebook
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
library(readxl)
#
library(gapminder)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


### 1. R Markdown
> (2 marks)
>
> Using the {gapminder} package, make a plot showing how the relationship between per capita GDP and life expectancy in 2007.


```{r}
gapminder

gapminder %>%
    filter(year == 2007) %>%
ggplot(mapping =  aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
  

```

```{r}
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap))
  
```

