---
title: "BIDA302 - assignment #3"
subtitle: "solutions"
output: html_document
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
#
library(gapminder)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> All of the packages and data you will need for this assignment is in the rstudio.cloud project. There is no need for you to download and install anything.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


## 1. Plotting
> (2 marks)
>
> Using the {gapminder} package, make a plot showing how the relationship between per capita GDP and life expectancy in 2007.


```{r}
# solution
gapminder

gapminder %>%
    filter(year == 2007) %>%
ggplot(mapping =  aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
  

```

```{r}
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(lifeExp))
  
```



## 2. Plotting
> (2 marks)
>
> This is a more complex extension of the {gapminder} plots you did as a hands-on exercise.
>
> Using your data manipulation skills, write a pipe that first calculates the median life expectancy for each continent for each year between 1952 and 2007, then plot a line graph showing those changes.



```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarise(lifeExp_median = median(lifeExp)) %>%
  #
  ggplot(mapping =  aes(x = year, y = lifeExp_median, colour = continent)) +
  geom_line()
  

```


## 3. Joins with HR data
> (8 marks)
> >
> 
> The Lahman baseball database contains a comprehensive listing of annual summaries for the major leagues of baseball, going back to 1871. 
> 
> For this question, smaller versions of three of the files have been created:
> 
> * "Master_2000s.csv" -- all active players from the years 2000 to 2018.
> 
> * "Salaries_2016.csv" -- the annual summary of each player's compensation, for the 2016 season.
> 
> * "Batting_2016.csv" -- the annual summary of each player's batting, for the 2016 season. (Note #1: that this omits pitchers who did not come to bat in the 2016 season. Note #2: players who were traded between teams mid-season show up with a record for their performance of each team.)
> 
> Think of these as the sorts of files that you would find in any human resources department: the Master file contains the personal details of each individual employed by the enterprise, 


### 3a. Read in the three files

```{r}
# solution
Master_2000s <- read_csv("Master_2000s.csv")

Batting_2016 <- read_csv("Batting_2016.csv")

Salaries_2016 <- read_csv("Salaries_2016.csv")

```

### 3b. Understanding the tables

> What is the key variable that links the three tables?
> 
> Which table has the primary key? What is the reason behind your answer?
> 
> Why would the designers of the table put only one variable in the Batting table that identifies a player?


### 3c. Joins

> Join the tables in such a way so that you can tabulate the number of batters that played on each team in 2016, by birth country.
>
> Note: because some of the non-key variables exist in multiple tables, {dplyr} will rename variables as `name.x` and `name.y` -- watch out for this! Take a look at your variables after you've completed the join.

```{r}
# solution
MLB_HR_data <- 
left_join(Batting_2016, Master_2000s, by = "playerID")


# alternate version, using a pipe
MLB_HR_data <- 
Batting_2016 %>%
  left_join(Master_2000s, by = "playerID") %>% 
  left_join(Salaries_2016, by = "playerID")

MLB_HR_data

MLB_HR_data %>% 
  group_by(teamID.x, birthCountry) %>% 
  tally() 


# NOTES
#
# * both tables have the variable "playerID"
# 
# * we want a `left_join`: all the observations from the active season, and adding the country of birth variable from "Master"


```


### 3d. Filtering joins

> Look at the number of records in `Salaries_2016` and `Batting_2016`. Earlier it was noted that the batting table _omits_ players ... but it's the longer of the two!
> 
> Use one of the filtering joins to see how many players are listed in `Batting_2016` that don't show up in `Salaries_2016`.


```{r}
# solution

semi_join(Batting_2016, Salaries_2016, key = "playerID")

semi_join(Batting_2016, Salaries_2016, key = "playerID") %>% 
  group_by(playerID) %>% 
  tally()

```


### 3e. Plot!

> Use your exploratory data analysis skills to create two visualizations to see what might explain player compensation (salary). First, compare it to the age of player, and then to the player's performance, using the crude measure of runs scored (the "R" variable in the batting table).
> 

```{r}
ggplot(MLB_HR_data, aes(x = birthYear, y = salary)) +
  geom_point()
```



```{r}
ggplot(MLB_HR_data, aes(x = R, y = salary)) +
  geom_point()
```



## REFERENCE

The source of the Lahman baseball data table is the R package [{Lahman}](https://cran.r-project.org/web/packages/Lahman/index.html), which makes more accessible the data from [Sean Lahman's historical baseball database](http://www.seanlahman.com/baseball-archive/statistics/). I modified the Batting and Master tables as follows:

```{r evaluate = FALSE}

library(Lahman)

Batting_2016 <- Batting %>%
  filter(yearID == 2016)
 
write_csv(Batting_2016, "Batting_2016.csv")
# ---
Master_2000s <- Master %>%
  filter(finalGame > "2000-01-01")
 
write_csv(Master_2000s, "Master_2000s.csv")
# ---
Salaries_2016 <- Salaries %>% 
 filter(yearID == 2016)
 
write_csv(Salaries_2016, "Salaries_2016.csv")



```







