---
title: "BIDA302 - assignment #3"
subtitle: "solutions"
output: html_document
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
library(readxl)
#
library(gapminder)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


### 1. Plotting
> (2 marks)
>
> Using the {gapminder} package, make a plot showing how the relationship between per capita GDP and life expectancy in 2007.


```{r}
gapminder

gapminder %>%
    filter(year == 2007) %>%
ggplot(mapping =  aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
  

```

```{r}
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(lifeExp))
  
```



### 2. Plotting

This is a more complex extension of the {gapminder} plots you did as a hands-on exercise.

Using your grouping and mutate skills, write a pipe that first calculates the median life expectancy for each continent for each year between 1952 and 2007, then plot a line graph showing those changes.



```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarise(lifeExp_median = median(lifeExp)) %>%
  #
  ggplot(mapping =  aes(x = year, y = lifeExp_median, colour = continent)) +
  geom_line()
  

```

