---
title: "Plotting in R"
subtitle: "hands-on examples, with answers"
output: html_notebook
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. -->

```{r setup}
library(tidyverse)

library(gapminder)
```

# 1. Plotting

## Objectives

* Produce scatter plots, bar graphs, and time series plots using {ggplot2}.

* Set universal plot settings.
 
* Describe what faceting is and apply faceting in a `ggplot()` call.
 
* Modify the aesthetics of an existing plot (including axis labels and color).
 
* Build complex and customized plots from data in a data frame.


## a short example


### 1. 
> (3 marks)
>
> a. Create a scatterplot using the {gapminder} data, showing all the countries in the world for 2007, the per capita GDP on the _x_ axis, and life expectancy on the _y_
>

 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

```

> b. Create a variant of the above plot, with each continent coloured differently

 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point()

```


> c. And now using multiple facets
 
```{r}
# solution

gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  facet_grid(~continent)

```

