---
title: "Plotting your data"
subtitle: "part of Exploratory Data Analysis"
author: "Martin Monkman"
date: "2019/07/04"
output:
  xaringan::moon_reader:
    css: ["default", "css_files/eigengrau.css", "css_files/eigengrau-fonts.css"]    
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
```


# Exploratory Data Analysis

"Exploratory data analysis is detective work." <br>(John W. Tukey)

--

>1. Generate questions about your data
>
>2. Search for answers by visualizing, transforming, and modeling your data
>
>3. Use what you learn to refine your questions and/or generate new questions


- Wickham and Grolemund, _R4DS_


??? 



---

class: inverse

# Anscombe's Quartet

### "A computer should make both calculations and graphs. Both sorts of outputs should be studied; each will contribute to understanding."

- F.J. Anscombe, "Graphs in Statistical Analysis", 1973



---
class: slide-banner

# The Quartet

.pull-left[

### Four data sets

- each set has 11 pairs of _x_ and _y_ values

]

.pull-right[

### Identical summary statistics:

- N = 11
- Mean of x = 9.0
- Mean of y = 7.5
- Multiple R2 = 0.667
- Regression equation: y = 3 + 0.5x
- And other statistical summary measures are also identical

]

---
class: slide-banner

# The Quartet

Four data sets with 11 pairs of _x_ and _y_ values]


All four have identical summary statistics:

| statistic   | value  |
| --- | ---: |
| N   | 11  |
| mean of _x_ | 9.0 |
| mean of _y_ | 7.5 |
| multiple R<sup>2</sup> | 0.667 |
| regression equation | _y_ = 3 + 0.5_x_ |

]


---
class: slide-banner


# The Exercise


### Read data into RStudio

### Calculate:

- Means of _x_ and _y_ values

- Multiple R<sup>2</sup>

### Visualize each set:

- Scatterplot (a.k.a. _x/y_ plot)

- Scatterplot with trend line and regression equation



---

# basic ggplot() call

`ggplot(<DATA>) + `

`  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)) `




---


`ggplot(anscombe_tidy) + 
  geom_point(mapping = aes(x = x, y = y)) `



```{r echo=FALSE, message=FALSE}
anscombe_tidy <- read_csv("anscombe_tidy.csv")
# example
ggplot(anscombe_tidy, aes(x = x, y = y)) + 
  geom_point() 

```
