<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  

The tables of functions are modified from originals by Gina Reynolds,[Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action)

-->

```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)
```

# Introduction to R and RStudio {#rstudio}


```{r echo = FALSE}

library(tidyverse)
library(gapminder)
```


For this course, we will be using the web browser implementation of R and RStudio, "rstudio.cloud". You can access this through any current web browser. Sign up for a free account here:  https://rstudio.cloud/plans/free

The link to the class space will be provided separately.



#### Packages

##### R Markdown

[R Markdown reference at RStudio](https://rmarkdown.rstudio.com/docs/)

[R Markdown cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf) (PDF file)


#### R Markdown

[Introduction to literate programming in R](https://remi-daigle.github.io/2017-CHONe-Data/Rmarkdown.nb.html) -- from Data Management Workshop à la CHONe





## Installing R on your own computer


If you decide to install R and RStudio on your own computer, follow the instructions provided here: [_R Cookbook, 2nd ed._](https://rc2e.com/gettingstarted#recipe-id001)

* 1.1 Downloading and Installing R

* 1.2 Installing RStudio

* 1.3 Starting RStudio



-30-