<!-- This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  

The tables of functions are modified from originals by Gina Reynolds,[Tidyverse in Action](https://github.com/EvaMaeRey/tidyverse_in_action) -->


# Exercise 1 - data wrangling {-}


## Setup {-}

This chunk of R code loads the packages that we will be using.

```{r setup_199, eval=FALSE}
#Package load

library(tidyverse)
library(stringr)

# data files
library(nycflights13)

```


## Introduction {-}

For this homework exercise, use the functions that were introduced in class to solve the following problems. **You shouldn't have to use any functions that are not covered in class!** [The course book](https://bookdown.org/martin_monkman/DataAnalyticsCodingFundamentals/) should be your first point of reference as you tackle the problems.



## Questions {-}

::: {.rmdtip}

### 1. Assign a variable  {-}

Why does this code chunk not work?

```
stringlenght <- 10

stringlength
```

Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)


<details>

<summary>
**Solution**
</summary>

First, the code chunk was missing the {r} portion.

Second, the object name is spelled differently in each case: it ends with `ht` in the first line, and `th` in the second.

</details>
:::


::: {.rmdtip}




### 2. Reducing the size of your data set  {-}

Starting with the {mpg} data frame, use `filter()` and `select()` to create a single table that contains _only_ the following records and variables:

* compact and subcompact Volkswagens from the model year 1999

* manufacturer, model, mileage in both city and highway 

<details>

<summary>
**Solution**
</summary>


First, view the data table:

```{r}
mpg

```

In this first solution, each evaluation gets its own `filter()` function:

```{r}
mpg |>
  filter(manufacturer == "volkswagen")  |> 
  filter(class == "compact" | class == "subcompact")  |>   ##  - use of "OR" `|` instead of `%in%`
  filter(year == 1999) |>
  select(manufacturer, model, cty, hwy)

```

Another possible solution combines the three evaluations into a single `filter()` function:


```{r}
mpg |>
  filter(manufacturer == "volkswagen",
         class %in% c("compact", "subcompact"),
         year == 1999) |>
  select(manufacturer, model, cty, hwy)

```

In this variant, the `filter()` function uses "&" instead of commas at the end of each evaluation:

```{r}

## variant
mpg |>
  filter(manufacturer == "volkswagen" &   # use of "&" instead of commas in filter
         class %in% c("compact", "subcompact") &
         year == 1999) |>
  select(manufacturer, model, cty, hwy)
```



</details>
:::


::: {.rmdtip}




***

### 3. Selecting variables  {-}

What happens if you include the name of a variable twice in a `select()` call? Demonstrate using the {gapminder} data.  (The package is installed in the workspace, but you will have to load it to access it...)

```{r}
library(gapminder)

```
<details>

<summary>
**Solution**
</summary>

The variable is _not_ duplicated. In the example below, the variable "year" is duplicated but it appears only once in the output from the `select()` function.

```{r}
gapminder |>
  select(country, year, lifeExp, year)

```


</details>
:::


::: {.rmdtip}

### 4. R Markdown  {-}

The following chunk of code produces a chart. When this notebook is previewed, the R code inside the chunk is also rendered.

Enter a chunk option so that the chart shows in the output, but the code does not.

(Hint: use the RMarkdown cheatsheet to find the solution!)


```{r}
# hide this code!
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```

<details>

<summary>
**Solution**
</summary>


The solution is to put the option "echo = FALSE" after the "r" in the opening bit of the code.


`{r, echo=FALSE}`
`ggplot(data = mpg) +`
`  geom_point(mapping = aes(x = displ, y = hwy))`



</details>
:::


::: {.rmdtip}

### 5. Flight data {-}

For Q3a-3d, use the data `nycflights13::flights`

> Sidebar: To see the names of all of the tables in the {nycflights13} package, use the `data()` function:

```{r}
data(package = "nycflights13")
```


Pay close attention to how the variables are coded. Type `?flights` into the console to access the details about this data in the "Help" pane.

Using the data in the `nycflights13::flights` table, find all flights that:

a. Had an arrival delay of two or more hours

b. Flew to Houston (there are two airports, with the codes IAH or HOU)

c. Were operated by United, American, or Delta (you may want to review the separate airlines table to check the codes for each airline)

d. Departed in summer (July, August, and September)

<details>

<summary>
**Solution**
</summary>




**Solution a**

_a. Had an arrival delay of two or more hours_

Note that variable `arr_delay` is measured in minutes, so we need to find the flights where that's greater than or equal to 120 minutes.

```{r}
filter(flights, arr_delay >= 120)
```



**Solution b**

_b. Flew to Houston (there are two airports, with the codes IAH or HOU)_


```{r}

filter(flights, dest == "IAH" | dest == "HOU")

```


Note that you could also use `%in%`

```{r}

filter(flights, dest %in% c("IAH", "HOU"))

# create named variable "houston"
houston <- c("IAH", "HOU")
filter(flights, dest %in% houston)

```


**Solution c**

_c. Were operated by United, American, or Delta_ 

This one takes a bit of effort--you have to find the codes that match the values in the variable `carrier` with the airline names.

In addition to the `flights` table, the {nycflights13} package has a number of other tables, including `airlines`

```{r}
nycflights13::airlines
```


From this, we can find:

* United == UA

* American == AA

* Delta == DL

```{r}
# the `%in%` approach -- other methods are possible!
filter(flights, carrier %in% c("AA", "DL", "UA"))

```



**Solution d**

d. Departed in summer (July, August, and September)

The variable `month` is numeric, so we have to use greater than and less than to define the range.

```{r}

filter(flights, month >= 7, month <= 9)

```


Using the `:` operator would also allow us to make a range. 

```{r}

filter(flights, month %in% 7:9)

```


You could also use "or" `|`, or `between()`

```{r}

filter(flights, month == 7 |
         month == 8 |
         month == 9)

filter(flights, between(month, 7, 9))

```


**Solution a-d**

It is possible to combine all four of the elements in Q6 into a single code sequence:

```{r}
# with each step as a separate `filter()` statement
flights |> 
  filter(arr_delay >= 120) |> 
  filter(dest == "IAH" | dest == "HOU") |> 
  filter(carrier %in% c("AA", "DL", "UA")) |> 
  filter(month %in% 7:9)

```

We could go one step further and combine it all into a single `filter()` function:

```{r}
flights |> 
  filter(arr_delay >= 120 &
           dest %in% c("IAH", "HOU") & 
           carrier %in% c("AA", "DL", "UA") & 
           month %in% 7:9)

```



</details>
:::



-30-
