<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  
-->


# Exporting data & graphics {#exporting}


```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)

```

## Saving data

R provides a variety of options for saving dataframes that have been created. For these examples, we will look at CSV and Excel files, but there are many other options.

### Writing a CSV file

Write the Canada records from gapminder as a CSV file. This example uses the `write_csv()` function that is within the {readr} package.

* ["Write a data frame to a delimited file"](https://readr.tidyverse.org/reference/write_delim.html), from the {readr} package site

```{r setup_571}
library(readr)
library(gapminder)

```

Filter so that only the records for Canada are included; assign to new object "gapminder_canada"

```{r}
gapminder_canada <- gapminder %>% 
  filter(country == "Canada")
  
```

Write the dataframe object as a csv file.

```{r}

write_csv(gapminder_canada, "gm_canada.csv")

```


## Saving graphs


