<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  
-->


# Exporting data & graphics {#exporting}


```{r echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE)
options(width = 100, dplyr.width = 100)

```

## Saving data

R provides a variety of options for saving dataframes that have been created. For these examples, we will look at CSV and Excel files, but there are many other options.

### Writing a CSV file

Write the Canada records from gapminder as a CSV file. This example uses the `write_csv()` function that is within the {readr} package.

* ["Write a data frame to a delimited file"](https://readr.tidyverse.org/reference/write_delim.html), from the {readr} package site

```{r setup_571}
library(readr)
library(gapminder)

```

Filter so that only the records for Canada are included; assign to new object "gapminder_canada"

```{r}
gapminder_canada <- gapminder %>% 
  filter(country == "Canada")
  
```

Write the dataframe object as a csv file.

```

write_csv(gapminder_canada, "gm_canada.csv")

```


## Writing an Excel file


{openxlsx}: Read, Write and Edit xlsx Files

* CRAN https://cran.r-project.org/web/packages/openxlsx/index.html

* package reference page: https://ycphs.github.io/openxlsx/


From the Introduction article at the package reference: https://ycphs.github.io/openxlsx/articles/Introduction.html





## Saving graphs

`ggsave()` (one of the functions in {ggplot2})

https://ggplot2.tidyverse.org/reference/ggsave.html


Let's make a plot of life expectancy changes in Canada, using the dataframe we made above:

```{r}

plot_gm_canada <- ggplot(gapminder_canada, aes(x = year, y = lifeExp)) +
  geom_line()

plot_gm_canada

```

To save this plot as a separate file, we can use the `ggsave()` function. Note that in this example, the image hasn't been saved into our environment as a plot object--the `ggsave()` function will save the last object that was created. In this version, we save the previously created plot as a png file.

```

ggsave("gapminder_canada_lifeexp.png")

```

The function has arguments that you might find useful: 

* specify the file type, 

* specify an object, and

* the change the dimensions to suit your purpose and data:

```

ggsave("gapminder_canada_lifeexp.jpg", 
  plot_gm_canada, 
  width = 9, height = 6)

```

