---
title: "pipes"
author: "Martin Monkman"
date: "2020/03/07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This exercise uses the SAFI data, and is a modification of content in the [Data Carpentries "R for Social Scientists"](https://datacarpentry.org/r-socialsci/)

Load the tidyverse and data frame (creates an object named `interviews`):

```{r, results = 'hide', purl = FALSE, message=FALSE}

## load the tidyverse
library(tidyverse)

interviews <- read_csv("data/SAFI_clean.csv", na = "NULL")

## inspect the data
interviews

## preview the data
# View(interviews)
```

We're going to focus on two of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions

### Why would you want to subset your data?

> ...it is rare that you get the data in exactly the right form you need. Often you’ll need to create some new variables or summaries, or maybe you just want to rename the variables or reorder the observations in order to make the data a little easier to work with.

([_R for Data Science_, Data transformation](https://r4ds.had.co.nz/transform.html))

## Selecting columns and filtering rows

To select columns of a
data frame, use `select()`. The first argument to this function is the data
frame (`interviews`), and the subsequent arguments are the columns to keep.

```{r, results = 'hide', purl = FALSE}
select(interviews, village, no_membrs, years_liv)
```

To choose rows based on a specific criteria, use `filter()`:

```{r, purl = FALSE}
filter(interviews, village == "God")
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame `interviews2` and use
that as input to the next function, like this:

```{r, purl = FALSE}
interviews2 <- filter(interviews, village == "God")
interviews_god <- select(interviews2, no_membrs, years_liv)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl = FALSE}
interviews_god <- select(filter(interviews, village == "God"), no_membrs, years_liv)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take the
output of one function and send it directly to the next, which is useful when
you need to do many things to the same dataset. Pipes in R look like `%>%` and
are made available via the **`magrittr`** package, installed automatically with
**`dplyr`**. If you use RStudio, you can type the pipe with <kbd>Ctrl</kbd>
+ <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> +
<kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl = FALSE}
interviews %>%
    filter(village == "God") %>%
    select(no_membrs, years_liv)
```

In the above code, we use the pipe to send the `interviews` dataset first
through `filter()` to keep rows where `village` is "God", then through
`select()` to keep only the `no_membrs` and `years_liv` columns. Since `%>%`
takes the object on its left and passes it as the first argument to the function
on its right, we don't need to explicitly include the data frame as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we take the data frame `interviews`, *then* we `filter`
for rows with `village == "God"`, *then* we `select` columns `no_membrs` and `years_liv`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl = FALSE}
interviews_god <- interviews %>%
    filter(village == "God") %>%
    select(no_membrs, years_liv)

interviews_god
```

Note that the final data frame (`interviews_god`) is the leftmost part of this expression.


-30-
