---
title: "summarize"
author: "Martin Monkman"
date: "2020/03/07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This exercise uses the SAFI data, and is a modification of content in the [Data Carpentries "R for Social Scientists"](https://datacarpentry.org/r-socialsci/)

Load the tidyverse and data frame (creates an object named `interviews`):

```{r, results = 'hide', purl = FALSE, message=FALSE}

## load the tidyverse
library(tidyverse)

interviews <- read_csv("data/SAFI_clean.csv", na = "NULL")

## inspect the data
interviews

## preview the data
# View(interviews)
```


### Split-apply-combine data analysis and the summarize() function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of
the `group_by()` function.


#### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the average household size by village:

```{r, purl = FALSE}
interviews %>%
    group_by(village) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

You may also have noticed that the output from these calls doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` over data frame.

You can also group by multiple columns:

```{r, purl = FALSE}
interviews %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

When grouping both by `village` and `membr_assoc`, we see rows in our table for
respondents who did not specify whether they were a member of an irrigation
association. We can exclude those data from our table using a filter step.


```{r, purl = FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum household size for each village for each group
(members of an irrigation association vs not):

```{r, purl = FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs),
              min_membrs = min(no_membrs))
```

It is sometimes useful to rearrange the result of a query to inspect the values. For instance, we can sort on `min_membrs` to put the group
with the smallest household first:


```{r, purl = FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs), min_membrs = min(no_membrs)) %>%
    arrange(min_membrs)
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of minimum household size:

```{r, purl = FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs),
              min_membrs = min(no_membrs)) %>%
    arrange(desc(min_membrs))
```

#### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each village, we would do:

```{r, purl = FALSE}
interviews %>%
    count(village)
```

For convenience, `count()` provides the `sort` argument to get results in decreasing order:

```{r, purl = FALSE}
interviews %>%
    count(village, sort = TRUE)
```


-30-


-30-
