---
title: "BIDA302 - assignment #2"
subtitle: "solutions"
output: html_notebook
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
library(readxl)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


### 1. R Markdown
> (2 marks)
>
> Using the R Markdown quick reference, figure out how to:
>
> a. Add a footnote.
>
> b. Add a horizontal rule.
>
> c. Add a block quote.


**Answer:**

This is the answer



### 2. Importing data: csv
> (5 marks)
>
> Comma delimited files are one of the most common formats used for flat files. The tidyverse has a package, {readr}, which will automatically load with the function `library(tidyverse)`.
>
> The reference page for {readr} is here: https://readr.tidyverse.org/index.html
>
>
> a. Read in the file "cy-concluded-cases-5-yr-regional-report-dashboard-2014-to-2018.csv", which has the number of concluded court cases in B.C. by a variety of dimensions.
> 
> The file is sourced from here: 
> https://catalogue.data.gov.bc.ca/dataset/courts-concluded-cases-5-yr-regional-report-dashboard-2014-2018


**Answer:**

```{r}
# solution
court_cases <- read_csv("cy-concluded-cases-5-yr-regional-report-dashboard-2014-to-2018.csv")
```

> b. Read the file again, but specify that the region code should be a character.

```{r}
# solution
court_cases <- 
  read_csv("cy-concluded-cases-5-yr-regional-report-dashboard-2014-to-2018.csv",
                        col_types = cols(
#                          Year = col_date(format = "%Y"),
                          `Court Region Code` = col_character()
                        ))
```


>c. Now you've read it in, use your data manipulation skills to find out how many cases were completed in each court region in 2018. Sort the table so the busiest regions are at the top.


```{r}
# solution
court_cases %>%
  filter(Year == 2018) %>%
  group_by(`Court Region`) %>%
  summarise(total_cases = sum(Count)) %>%
  arrange(desc(total_cases))
```

> d. Create a pivot table showing the 2018 case completion, with the region as the rows and the court class as the columns. 

```{r}
# solution
court_cases %>%
  filter(Year == 2018) %>%
  group_by(`Court Region`, `Court Class`) %>%
  summarise(total_cases = sum(Count)) %>%
  pivot_wider(names_from = `Court Class`, values_from = total_cases)
```


### 3. Importing data: Excel
> (5 marks)
>
> Use the {readxl} package to read in the file "regional_district_monthly_2019_mod.xlsx" 
>
> This data is published by the Ministry of Finance, and shows a variety of statistics associated with the Property Transfer Tax.
>
> The original file has been modified:
> 
> * filter for the Capital Regional District (CRD) and
>
> * the resulting table is transposed, so the rows now become the columns.
>
> The website source for the data is here: https://catalogue.data.gov.bc.ca/dataset/property-transfer-tax-data-2019
>
> The reference page for {readxl} is here: https://readxl.tidyverse.org/
>
>
>a. What are the names of the sheets in this file?

```{r}
# solution

excel_sheets("regional_district_monthly_2019_mod.xlsx")

```

> b. Read in the contents of the sheet with the shortest name.

```{r}

read_excel("regional_district_monthly_2019.xlsx", sheet = "CRD_MONTHLY_2019")

```

c. What issues are there with the rows? How is this impacting how the values in the columns are being represented?

d. Change the options for your read function to address this.

```{r}
# solution

read_excel("regional_district_monthly_2019.xlsx", sheet = "CRD_MONTHLY_2019", 
           skip = 2)

```

> 
> e. Use your data manipulation skills to calculate the number of residential property transfers where there was foreign > involvement.
> 
> **Note:** Depending on what you did in 3.d, you may need to go back and read the data in again, adding more options.


```{r}
# solution

read_excel("regional_district_monthly_2019.xlsx", sheet = "CRD_MONTHLY_2019", 
           skip = 2,
           na = "nr")

```

