---
title: "Variable Types"
subtitle: "in R"
author: "BIDA302 - lesson 2"
date: "2019/11/09"
output:
  xaringan::moon_reader:
    css: [/static/css_files/eigengrau.css, /static/css_files/eigengrau-fonts.css]    
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# Martin Monkman
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

# to knit:
# see https://github.com/yihui/xaringan/issues/29

# point up-directory to css and img folders

#xaringan::inf_mr('lessons/class_02/slides_2_1_variable_types.Rmd')

```


```{r}
mpg
```


---

class: inverse

# Variable types 1


`int` stands for integers.

`dbl` stands for doubles, or real numbers.

`chr` stands for character vectors, or strings.

`date` stands for dates.

`dttm` stands for date-times (a date + a time).

---

class: inverse

# Variable types 2 

`lgl` stands for logical, vectors that contain only TRUE or FALSE.

`fctr` stands for factors, which R uses to represent categorical variables with fixed possible values.

`NA` are explicitly missing values.


---

# Missing values `NA`

* missing values are contagious -- one in the input will return an `NA` in the output

A simple example:

```{r}

1 + NA

```



---

Or a longer example...

`displ_class_by_cyl`

```{r echo=FALSE}
# summary table - class by cylinder
displ_class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) %>%
  arrange(desc(displ_mean)) %>%
  pivot_wider(names_from = cyl, values_from = displ_mean) %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

---

```{r}
mean(displ_class_by_cyl$displ_mean)
```

--

`na.rm` removes `NA`s from the calculation

```{r}
mean(displ_class_by_cyl$displ_mean, na.rm = TRUE)
```





---

## Functions for missing values

Dealing with those pesky `NA` values  

| function | action |
| : -----  | ---: |
| `is.na(x)` | returns TRUE or FALSE for each value in `x` |
| `anyNA(x)` | returns a single TRUE or FALSE |
| `na.rm = TRUE` | remove `NA` values when running function|



```{r}
x <- c(1, NA, 3)

is.na(x)

```



