---
title: "create correlated values"
author: "Martin Monkman"
date: "31/03/2021"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial source

https://stat.ethz.ch/pipermail/r-help/2007-April/128925.html

https://stats.stackexchange.com/questions/83172/generate-two-variables-with-precise-pre-specified-correlation
>With one variable we could subtract the data mean and divide by the standard deviation to get a variable with mean 0 and sd 1, then multiply by the target sd and add the target mean. Here we do the same thing, but use the "square root" of the covariance/correlation matrix, first to make sure the covariance matrix is the identity, then to force the target correlations. – Greg Snow Nov 27 '18 at 17:41


```{r setup}
library('MASS')
# tidyverse
library(tidyverse)
library(glue)
```


```{r}
samples = 200
#r = 0.95
r = sqrt(0.5)
r

data = mvrnorm(n=samples, mu=c(0, 0), Sigma=matrix(c(1, r, r, 1), nrow=2), empirical=TRUE)
X = data[, 1]  # standard normal (mu=0, sd=1)
Y = data[, 2]  # standard normal (mu=0, sd=1)

cor(X, Y)  # yay!
cor(X*0.01 + 42, Y*3 - 1)  # Linear transformations of X and Y won't change r.
```


```{r}

df_data <- as_tibble(cbind(X, Y))

summary(lm(Y ~ X, df_data))

ggplot(df_data, aes(x = X, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm")
```


```{r}
# create the initial x variable
x1 <- rnorm(100, 15, 5)

# x2, x3, and x4 in a matrix, these will be modified to meet thecriteria
x234 <- scale(matrix( rnorm(300), ncol=3 ))

# put all into 1 matrix for simplicity
x1234 <- cbind(scale(x1),x234)

# find the current correlation matrix
c1 <- var(x1234)

# cholesky decomposition to get independence
chol1 <- solve(chol(c1))

newx <-  x1234 %*% chol1 

# check that we have independence and x1 unchanged
zapsmall(cor(newx))
all.equal( x1234[,1], newx[,1] )

# create new correlation structure (zeros can be replaced with other rvals)
newc <- matrix( 
c(1  , 0.4, 0.5, 0.6, 
  0.4, 1  , 0  , 0  ,
  0.5, 0  , 1  , 0  ,
  0.6, 0  , 0  , 1  ), ncol=4 )

# check that it is positive definite
eigen(newc)

chol2 <- chol(newc)

finalx <- newx %*% chol2 * sd(x1) + mean(x1)

# verify success
mean(x1)
colMeans(finalx)

sd(x1)
apply(finalx, 2, sd)

zapsmall(cor(finalx))
pairs(finalx)

all.equal(x1, finalx[,1])


```

```{r}
# function

samples <- 10
r_square <- 0.9

correlated <- function(samples, r_square){
  r = sqrt(r_square)
  data <- mvrnorm(n=samples, 
                 mu=c(0, 0), 
                 Sigma=matrix(c(1, r, r, 1), nrow=2), 
                 empirical=TRUE)
  data <- as_tibble(data) %>% 
    rename(X = V1,
           Y = V2)
  
}


```


```{r}

sample <- 200
r_squared <- 0.9

df_data <- correlated(sample, r_squared)

summary(lm(Y ~ X, df_data))

ggplot(df_data, aes(x = X, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = glue('n = {sample}, R-squared = {r_squared}'))

```

