---
title: "Iteration"
author: "Martin Monkman"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---
```{r}

library(tidyverse)
#
library(purrr)
#library(tidyr)
#
library(gapminder)

```


## Iteration

Like functions, iteration is a way to reduce repetion in your code, and repetition in copy-and-paste tasks.

Iteration:  

> helps you when you need to do the same thing to multiple inputs: repeating the same operation on different columns, or on different datasets. ([_R4DS_, "21 Iteration"](https://r4ds.had.co.nz/iteration.html))


A simple example, with 4 columns of 50 records each, with a normal distribution centred on 100. (This is a variation on what's in _R4DS_.)


```{r}
set.seed(8675309)

df <- tibble(
  a = rnorm(50, mean = 100, sd = 10),
  b = rnorm(50, mean = 100, sd = 10),
  c = rnorm(50, mean = 100, sd = 10),
  d = rnorm(50, mean = 100, sd = 10)
)
  
df

```

Calculate the median of each:

```{r}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

Or, write a for loop:

```{r}
# define the output object before you start the loop --
#   this makes your loop more efficient
output <- vector("double", ncol(df))  # 1. output

# the loop
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}

# print the output object
output

```


---

## Detour: square brackets!

These are _accessors_ -- a method of accessing data by defining the location of that data in the object.

Square brackets inside square brackets is old-school way of referencing a particular location in a list.

```{r}

# one square bracket: the first list in the object
#  - think of this a subset of the object
df[1]

# two square brackets: the contents of the first object
df[[1]]

# two values inside two square brackets: this is a single value, [[row, column]]
df[[50, 1]]



```

For a good explanation of R's accessors and the application in selecting vector and matrix elements, see 

* ["\[, \[\[, $: R accessors explained"](https://www.r-bloggers.com/r-accessors-explained/)) by Christopher Brown

* [_R Cookbook_, 2nd ed., "2.9 Selecting Vector Elements"](https://rc2e.com/somebasics#recipe-id039) by JD Long and Paul Teetor


---

## Data frame iteration

Let's calculate the mean of every column in the `mtcars` data frame:

```{r}

mtcars

```



Advice from _R4DS_:
> Think about the output, sequence, and body **before** you start writing the loop.


* Note that we can use the `names()` function to pull the variable names (column names) from the data table.

```{r}
# create the output vector
output <- vector("double", ncol(mtcars))

# the names() function 
names(output)
# assign the names to the `output` vector
names(output) <- names(mtcars)

# the loop
for (i in names(mtcars)) {
  output[i] <- mean(mtcars[[i]])
}

# print the output
output

```


## A very practical case

(This is modified from ["21.3.5 Exercises"](https://r4ds.had.co.nz/iteration.html#exercises-56) in _R for Data Science_, which uses CSV files.)


Imagine you have a directory full of Excel files that you want to read in. You have their paths in a vector, files <- dir("data/", pattern = "\\.xls*$", full.names = TRUE), and now want to read each one with `read_excel()`. Write the for loop that will load them into a single data frame.


First, let's get the names of those files that are in the sub-folder "data":

```{r}

files <- dir("data/", pattern = "\\.xls*", full.names = TRUE)
files

```

We now know find out how many files there are, and can use that to assign our object:

```{r}

df_list <- vector("list", length(files))

```

```{r}
for (i in seq_along(files)) {
  df_list[[i]] <- readxl::read_excel(files[[i]])
}
```

This creates an interesting R object--it's a list of data frames! Let's take a look:

```{r}
df_list
```

The last step is to use the function `bind_rows()` to combine the list of data frames into a single data frame.

```{r}

df <- bind_rows(df_list)

df 

```





## Resources

[_R4DS_, "21 Iteration"](https://r4ds.had.co.nz/iteration.html)

JD Long and Paul Teetor, [R Cookbook, 2nd ed., "Iterating with a Loop"](https://rc2e.com/simpleprogramming#recipe-for_loop), 2019

