---
title: "Transform Data"
subtitle: "hands-on examples, with answers"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the original work at https://github.com/rstudio/master-the-tidyverse by RStudio and https://github.com/cwickham/data-science-in-tidyverse-solutions. -->

```{r setup}
library(tidyverse)
library(gapminder)

```

# Transform Data 1: manipulate cases

This hands-on exercise uses functions from the {dplyr} package:




## gapminder

These exercises use a summary of data about the countries of the world, drawn from the gapminder.org tables. This table is stored in the R package {gapminder}, which is already loaded.

To view the data in the console or in the R Markdown output, enter the name of the data table.

```{r}
gapminder
```



## Choosing rows with `filter()` 


## Your Turn 1.2: filter() rows

See if you can use the logical operators to manipulate our code below to show:

The data for Canada

```{r}
# example
filter(.data = gapminder, country == "New Zealand")

# example
filter(gapminder, country == "New Zealand")

# answer
filter(gapminder, country == "Canada")

```

All data for countries in Oceania

```{r}
# example
filter(gapminder, country == "New Zealand")

# answer
filter(gapminder, continent == "Oceania")
```

Rows where the life expectancy is greater than 82
```{r}
# example
filter(gapminder, country == "New Zealand")

# answer
filter(gapminder, lifeExp > 82)
```


## Your Turn 1.3: multiple criteria

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r}
# example 
gapminder %>%                                       # Note use of `%>%` pipe symbol
  filter(country == "New Zealand", year > 2000)

# answer
gapminder %>%
  filter(country == "United States", year < 1970)

# alternative answer
gapminder %>%
  filter(country == "United States" & year < 1970)  # Note use of `&` rather than comma

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r}
# answer
gapminder %>%
  filter(year == 2007, lifeExp < 50 | lifeExp > 75) # Note that you have to state `lifeExp` twice, as
                                                    # there are two comparisons.
                                                    # What happens if you replace the comma after 2007 with `&`?
                                                    # - Why do you think this might be happening?



```



* Create a list object and return records for any of "New Zealand", "Canada" or "United States"


```{r}
# answer
gapminder %>%
  filter(country %in% c("New Zealand", "Canada", "United States"))
  
# alternate answer
country_list <- c("New Zealand", "Canada", "United States")   # create object that is list of countries
gapminder %>%
  filter(country %in% country_list)                           # filter uses that object 
  

```

* Return records for all continents except Europe


```{r}
# answer
gapminder %>%
  filter(continent != "Europe")
  
# answer with a double-check
gapminder %>%
  filter(continent != "Europe") %>%
  distinct(continent)                 # the `distinct()` function returns the list of all the 
                                      # continent names after the filter is applied

```

## Your Turn 1.4: sorting with `arrange()`

Find the records with the smallest population.
```{r}
#answer
gapminder %>%
  arrange(pop)

```

Find the records with the largest GDP per capita.
```{r}
#answer
gapminder %>% arrange(desc(gdpPercap))

#answer
arrange(gapminder, desc(gdpPercap))

#answer
gapminder %>% arrange(-gdpPercap)


```


## Your Turn 1.5: creating new variables

Calculate a new variable with total GDP

```{r, error = TRUE}
# answer
gapminder %>%
  mutate(totalGDP = pop * gdpPercap)


```

Create a summary table with the population and GDP by continent for the year 1952  

```{r}
# answer
gapminder %>%
  filter(year == 1952) %>%               # filter to get the records we need
  mutate(totalGDP = pop * gdpPercap) %>% # note that we need to add the new variable calculated above!
  group_by(continent) %>%                # this defines the grouping category
  summarize(pop = sum(pop),              # 
            totalGDP = sum(totalGDP))    #
```




## Your Turn 1.6

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: use "Transformation" cheatsheet, you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?

```{r}
# answer
gapminder %>%
  mutate(prev_lifeExp = lag(lifeExp))

# challenge hint: 
# - note that there is an `NA` in the first row, since there's no prior row to pull from
# - look at the `prev_lifeExp` when the country changes from Afghanistan to Albania 
```


## Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  

***

# Transform Data 2: summarize

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->


```{r}
library(tidyverse)
library(gapminder)

```

## gapminder

Take a quick look at the data in the {gapminder} package.

```{r}
gapminder
```



## Your Turn 2.1

Use summarise() to compute three statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of unique countries

```{r}
# answer
gapminder %>%
  summarize(year_min = min(year),
            year_max = max(year),
            n_countries = n_distinct(country))

```

## Your Turn 2.2

Extract the rows for African countries in 2007.

Then find:

1. The number of unique countries
2. The median life expectancy of African countries as a group

```{r}
# answer
gapminder %>% 
  filter(continent == "Africa", year == 2007) %>% 
  summarise(n_countries = n_distinct(country),
    med_life_exp = median(lifeExp))


```


## Your Turn 2.3

Find the median life expectancy by continent in 2007.

```{r}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>% 
  summarise(med_life_exp = median(lifeExp)) 

```


## Your Turn 2.4

Find the median life expectancy by continent for each year in the dataframe.

```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp))

```
## Your Turn 2.5

Brainstorm with your neighbor the sequence of operations to find:  the country with biggest jump in life expectancy  (between any two consecutive records) for each continent.

## Your Turn 2.6

Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r}
# One of many solutions
gapminder %>%
  group_by(country) %>%
  mutate(prev_lifeExp = lag(lifeExp),
         jump = lifeExp - prev_lifeExp) %>%
  arrange(desc(jump))


```


## Take aways

* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`

***



# Transform Data 3: pivot

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->


```{r}
library(tidyverse)
# data packages
library(gapminder)

```


## Your Turn 3.1

A pivot table crosstab using the {mpg} data package of automobile fuel economy.

First, a look at the source table.

```{r}
mpg
```

Using `group_by` and `summarise` to create a summary table of the average engine displacement by vehicle class:

```{r}
mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```

Now do the same with vehicle class and number of cylinders:

```{r}
# answer
class_by_cyl <- mpg %>%                 # note that an object `class_by_cyl` is created
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl                            # in order to print the object, just name it

```


What happens if you reverse the order of the grouping variables?



Create a pivot (crosstab) table with class by number of cylinders:


```{r}
# answer
# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot


```

Now, unpivot it back to the original structure:

```{r}
# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

What do you notice about the structure of the unpivoted table?



## Your Turn 3.2

Making some summary tables with the {gapminder} data.

Create a table where:

* object name `lifeExp_by_continent` that has 
* the median life expectency for each continent for every year in the dataframe, 
* where the continent is the column and the year is the row.

(This is a restructuring of the table from 2.4)

```{r}
# answer

lifeExp_by_continent <- gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp)) %>%
  pivot_wider(names_from = continent, values_from = median_lifeExp)

lifeExp_by_continent

```

## Your Turn 3.3

Create a table that shows:

* GDP per capita for
* the countries Canada, United States, and Mexico are the columns, and 
* the years after 1980 are the rows

```{r}

gapminder %>%
  filter(year > 1980,
         country %in% c("Canada", "United States", "Mexico")) %>%
  select(country, year, gdpPercap) %>%        # select() is covered in section 4
  pivot_wider(names_from = country, values_from = gdpPercap)

```



## Take aways

* Restructure data tabulations with `pivot_wider()` and `pivot_longer`  

* Reinforce other data transformation functions


<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->


```{r}
library(tidyverse)
# data packages
library(gapminder)

```


***

## Choosing columns


Using `select()` from the package {dplyr} to reduce the number of variables:

```{r}
# answer
gapminder %>%                                
  select(country, year, lifeExp) 

```



You can use the minus sign to drop variable--that is, to define ones to _not_ select. 

In this example, note we can use `c()` to create a list:


```{r}
# answer
gapminder %>%                                
  select(-(c(continent, pop, gdpPercap))) 

# alternate answer
gapminder %>%                                
  select(-continent, -pop, -gdpPercap) 

```



You can define a range of columns to select by using the `:`

```{r}
# answer
gapminder %>%
  select(country, year:pop)
```


And moving columns is accomplished with the function `relocate()`. 

By specifying a column, it gets moved to the first (left-most) column. In this example, we can move "continent" to the first spot.

```{r}
# example

gapminder %>% 
  relocate(continent)

```




## Your turn

(add `select()` example)




How would you move "lifeExp" to the last (furthest right) column?

```{r}
# solution: specify all the other columns to move ahead of it

gapminder %>% 
  relocate(country, continent, year, pop, gdpPercap)

# solution: specify the one you don't want to move

gapminder %>% 
  relocate(-(lifeExp))

# solution: use `.after`

gapminder %>% 
  relocate(lifeExp, .after = gdpPercap)

```



## Take aways

* Restructure data frame columns with `select()` and `relocate()`





