---
title: "Transform Data 2: summarize"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->


```{r setup}
library(tidyverse)
library(gapminder)

```

> Please refer to [_R for Data Science_, Data transformation](https://r4ds.had.co.nz/transform.html) for more information about this section.

When working with data, one of the first things in our analysis is often the calculation of summary statistics. We may want the number of cases, or the maximum value, or the mean of all the records. 

Here's a list of the {dplyr} functions that you'll be using:


|function |action |
| :--- | :--- |
| `group_by()`  | *declare subsets in data*|
| `summarize()` | *summarize the data, by groups if they have been declared*|
| `tally()`     | *counting (by groups if group_by() applied)*|




## gapminder

```{r}
gapminder
```


## Your Turn 2.1

Use summarise() to compute two statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of unique countries

```{r}


```

## Your Turn 2.2

Extract the rows for African countries in 2007.

Then find:

1. The number of unique countries
2. The median life expectancy of African countries as a group

```{r}


```


## Your Turn 2.3

Find the median life expectancy by continent in 2007.

```{r}

```


## Your Turn 2.4

Find the median life expectancy by continent for each year in the dataframe.

```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp))

```
## Your Turn 2.5

Brainstorm with your neighbor the sequence of operations to find:  the country with biggest jump in life expectancy  (between any two consecutive records) for each continent.

## Your Turn 2.6

Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r}


```


***

# Take aways

* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`
