---
title: "Transform Data 1: manipulate cases"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(gapminder)

```

## Objectives

At the end of this exercise, you will demonstrate the use of the following functions from the {dplyr} package:


|function |action |
| :--- | :--- |
| `filter()`    | *keep rows*|
| `select()`    | *keep variables (or drop them -var)*|
| `mutate()`    | *create a new variable*|
| `arrange()`    | *sort from smallest to largest*|
| `arrange(desc())`    | *sort from largest to smallest*|



***

> Please refer to [_R for Data Science_, Data transformation](https://r4ds.had.co.nz/transform.html) for more information about this section.

When working with data, particularly large data sets, you encounter a situation where you need to: 

* subset the data so that it contains only those _records_ (rows) you are interested in (Why deal with a data table that has every province in Canada when you just need the British Columbia data?)

* subset the data so that it contains only those _variables_ (columns) you are interested in (Why keep looking at variables that do not factor into your analysis, or you will not print in your report?)

* create new variables, often through calculations based on variables in your data



To achieve these three goals, we will be using functions from the tidyverse package {dplyr}. The functions are verbs that describe what they do. Here's a list of the {dplyr} functions that you'll be using:


|function |action |
| :--- | :--- |
| `filter()`    | *keep rows*|
| `select()`    | *keep variables (or drop them -var)*|
| `mutate()`    | *create a new variable*|
| `arrange()`    | *sort from smallest to largest*|
| `arrange(desc())`    | *sort from largest to smallest*|



### other operators

In addition to these functions, R has a number of operators that add significantly to your code.

**Arithmetic operators**

| Operator | Description |
| :--- | :--- |
| + | *plus* |
| - | *minus* |
| \* | *multiplication*  |
| / | *division* |
| ^ | *exponential* |

<P>

**Boolean operators** are a special type of operator that return TRUE or FALSE.  

| Operator | Description |
| :--- | :--- |
| ==    | *equal, tests equality* |
| !=    | *not equal, tests inequality* |
| \>    | *greater than, tests greater than* (also >=) |
| <     | *less than, tests less than* (also <=) |
| %in%  | *contains, tests inclusion* |

<P>

Boolean operators can be combined with *and* or *or*.  

| Operator | Description |
| :--- | :--- |
| `&`     | *and, returns true if preceding and following are both true, else false* |
| &#124;  | *or, returns true if either preceding and following are true, else false* |


* see [_R for Data Science_, "Comparisions""](https://r4ds.had.co.nz/transform.html#comparisons) and ["Logical Operators"](https://r4ds.had.co.nz/transform.html#logical-operators).]





## gapminder

These exercises use a summary of data about the countries of the world, drawn from the gapminder.org tables. This table is stored in the R package {gapminder}.

To view the data in the console or in the R Markdown output, enter the name of the data table:

```{r}
gapminder
```

## Your Turn 1: filter() rows

The `filter()` function allows us to specify the observations (records) that we want to keep in our data analysis.

The general structure of a `filter()` function is:

`filter(.data, variable <test> value)`

Note that whatever comes first in the `filter()` function will be read as the data frame.


See if you can use the logical operators to manipulate our code below to show:

* the data for Canada

```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania

```{r}
filter(gapminder, country == "New Zealand")
```


Rows where the life expectancy is greater than 82
```{r}
filter(gapminder, country == "New Zealand")
```


Return records for all continents except Europe


```{r}

```



## Your Turn 2: multiple criteria


### Introducing the pipe operator ` %>% `

(see [_R for Data Science_, Pipes](https://r4ds.had.co.nz/pipes.html) for a deep dive into pipes)

The pipe operator ` %>% ` allows us to put the data first, and then have other functions work with the outcome of the first line.

For example, we could write

`function(data, arg)` or, with a pipe

`data %>% function(arg)` 

If we want to do anything with the first statement, we have to assign it to an intermediate object, and then use that object in the second statement. With the pipe, we can simply add another pipe and the results of the first function are passed to the second statement.

As we get to increasingly complex statements, the power of this will become more evident.

***

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r filter4_example, exercise=TRUE, exercise.eval=FALSE}
# example 
gapminder %>%                                       # Note use of `%>%` pipe symbol
  filter(country == "New Zealand", year > 2000)

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r}


```

### Creating lists

R allows us to create lists using the `c()` function.

A list can be referenced in a `filter()` statement using the `%in%` as the evaluator.

In this example, we filter to get three countries, "New Zealand", "Canada", and "United States".

Note that to get all three, the statement is _OR_, not _AND_! There are no cases where `country == "New Zealand"` and simultaneously `country == "Canada"` ... 

```{r}

gapminder %>% 
  filter(country == "New Zealand" |
           country == "Canada" |
           country == "United States")

```

Instead of this rather verbose syntax, we can achieve the same result by creating a list. 


```{r filter6_example-solution}
# answer
gapminder %>%
  filter(country %in% c("New Zealand", "Canada", "United States"))

```

Another option would be to create a named object that contains our list, and then use the named object in our `filter()`


```{r}
  
# alternate answer
country_list <- c("New Zealand", "Canada", "United States")   # create object that is list of countries

gapminder %>%
  filter(country %in% country_list)                           # filter uses that object 
  

```


* Create a list object and return records for any of "Belguim", "India", "Denmark" or "Angola" 


```{r}

```


* Return the records for all continents except Europe

```{r filter7_example}

```


```{r filter7_example-solution}
# answer
gapminder %>%
  filter(continent != "Europe")
  
# answer with a double-check
gapminder %>%
  filter(continent != "Europe") %>%
  distinct(continent)                 # the `distinct()` function returns the list of all the 
                                      # continent names after the filter is applied

```


## Your Turn 3: sorting

Sort the records so that those with the smallest population are shown first.

```{r}

```

Sort the records so that those with the largest GDP per capita are shown first.
```{r}

```


## Your Turn 4: creating new variables


Use the `mutate()` function to calculate a new variable with total GDP, by multiplying the total population by the GDP per capita.

```{r, error = TRUE}



```

Create a summary table with the population and GDP by continent for the year 2007  

```{r}

```

## Your Turn 5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: use "Transformation" cheatsheet, you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?

```{r}
gapminder %>%
  mutate()
```




## Your Turn 6: select cases

{gapminder} again

```{r}
gapminder
```

Using `select()` to reduce the number of variables:

```{r}
gapminder %>%                                
  select(country, year, lifeExp) 

```



You can use the minus sign to drop variable--that is, to define ones to _not_ select. 

In this example, note we have to use `c()` to create a list:


```{r}
gapminder %>%                                
  select(-(c(continent, pop, gdpPercap))) 

```



You can define a range of columns to select by using the `:`

```{r}
gapminder %>%
  select(country, year:pop)
```



***

# Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  
* Restructure data frames with `select()`  

