---
title: "Transform Data 2: summarize"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->


```{r setup}
library(tidyverse)
library(gapminder)

```

## gapminder

```{r}
gapminder
```


## Your Turn 2.1

Use summarise() to compute two statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of unique countries

```{r}


```

## Your Turn 2.2

Extract the rows for African countries in 2007.

Then find:

1. The number of unique countries
2. The median life expectancy of African countries as a group

```{r}


```


## Your Turn 2.3

Find the median life expectancy by continent in 2007.

```{r}

```


## Your Turn 2.4

Find the median life expectancy by continent for each year in the dataframe.

```{r}

gapminder %>%
  group_by(continent, year) %>%
  summarize(median_lifeExp = median(lifeExp))

```
## Your Turn 2.5

Brainstorm with your neighbor the sequence of operations to find:  the country with biggest jump in life expectancy  (between any two consecutive records) for each continent.

## Your Turn 2.6

Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r}


```


***

# Take aways

* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`
