---
title: "Transform Data 1: manipulate cases"
output: html_notebook
---

<!-- This file by Charlotte Wickham (with some modifications by Martin Monkman) is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(gapminder)

```

## gapminder

```{r}
gapminder
```

## Your Turn 1: filter() rows

See if you can use the logical operators to manipulate our code below to show:

The data for Canada

```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania

```{r}
filter(gapminder, country == "New Zealand")
```

Rows where the life expectancy is greater than 82
```{r}
filter(gapminder, country == "New Zealand")
```


## Your Turn 2: multiple criteria

Use Boolean operators to alter the code below to return only the rows that contain:

* United States before 1970

```{r}

gapminder %>%
  filter(country == "New Zealand", 
         year > 2000)

```

*  Countries where life expectancy in 2007 is below 50 or over 75

```{r}


```



* Create a list object and return records for any of "New Zealand", "Canada" or "United States"


```{r}

```

* Return records for all continents except Europe


```{r}

```

## Your Turn 3: sorting

Find the records with the smallest population.
```{r}

```

Find the records with the largest GDP per capita.
```{r}

```


## Your Turn 4: creating new variables

Calculate a new variable with total GDP

```{r, error = TRUE}



```

Create a summary table with the population and GDP by continent for the year 2007  

```{r}

```

## Your Turn 5

Alter the code to add a `prev_lifeExp` column that contains the life expectancy from the previous record.

(Hint: use "Transformation" cheatsheet, you want to offset elements by one)

Extra challenge: Why isn't this quite the 'life expectency five years ago'?

```{r}
gapminder %>%
  mutate()
```


***

# Take aways

* Extract cases with `filter()`  
* Filter using a list with `%in%`
* Make new variables, with `mutate()`  
* Connect operations with `%>%`  
