---
title: "lesson_1_02"
author: "Martin Monkman"
date: "11/10/2019"
output: html_notebook
---

<!-- This file by Martin is licensed under a Creative Commons Attribution 4.0 International License https://creativecommons.org/licenses/by/4.0/, adapted from  -->
<!-- * orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->
<!-- * orignal work at https://r4ds.had.co.nz/ by Garrett Grolemund and Hadley Wickham. -->
<!-- * orignal work at https://github.com/cwickham/data-science-in-tidyverse and https://github.com/cwickham/r_intro_bc_stats by Charlotte Wickham. -->



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Load the required packages

For this exercise, we will load the packages within the tidyverse

```{r}
library(ggplot2)
```


Built into Base R and other packages are data sets--handy for learning R (before we get to reading and writing data files)!

So let's use the `mpg` data frame (we'll talk more about data structures later). The data is the fuel consumption of 234 different cars, collected by the U.S. Environmental Protection Agency. Getting the data is simple:

```{r}

mpg

```

What happens if you type `View(mpg)` into the console?

***

We can create an x-y scatter plot using R's {ggplot2} package:

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

```


