---
title: "BIDA302 - assignment #1"
output: html_notebook
---

```{r}

library(tidyverse)
# data files
library(nycflights13)

```


## Introduction

"code through"

- example blogs


## Questions


### 1. Why does this code not work?

```{r}
stringlenght <- 10

stringlength
```


Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)

(from [_R for Data Science_, "Workflow Basics"](https://r4ds.had.co.nz/workflow-basics.html#practice))



### 2. Reducing the size of your data set

Starting with the {mpg} data frame, use `filter()` and `select()` to create a shorter form of this table that contains the following records and variables:

* Volkswagen compact and subcompact models from the model year 1999

* manufacturer, model, mileage in both city and highway 

```{r}

mpg


```



***

### 3. Car mileage - pivot table

In Lesson 1, we used the {mpg} package to calculate the average displacement by class of car:

```{r}
mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```


We then created a pivot (crosstab) table with class by number of cylinders, and then unpivoted that table:

```{r}
class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) %>%
  arrange(desc(displ_mean))

# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot

# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

What happens when you repeat the first chunk of code to calculate the average displacement of each class of cars but with the new object `mpg_class_by_cyl`?

```{r}

displ_class_by_cyl %>%
  group_by(class) %>%
  summarise(mean = mean(displ_mean))

```

Solve this problem so that the table looks like the first one in this question. Use any and all of the reference materials at your disposal!

***


### 4. Flight data

For Q3a-3d, use the data `nycflights13::flights`

Find all flights that:

a. Had an arrival delay of two or more hours
b. Flew to Houston (IAH or HOU)
c. Were operated by United, American, or Delta
d. Departed in summer (July, August, and September)


(from [R for Data Science, "Data transformation"](https://r4ds.had.co.nz/transform.html))

