---
title: "Transform Data 3: pivot"
output: html_notebook
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License -->


```{r setup}
library(tidyverse)
# data packages
library(gapminder)

```


## Your Turn 3.1

A pivot table crosstab using the {mpg} data package of automobile fuel economy.

First, a look at the source table.

```{r}
mpg
```

Using `group_by` and `summarise` to create a summary table of the average engine displacement by vehicle class:

```{r}
mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```

Now do the same with vehicle class and number of cylinders:

```{r}
class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) 

class_by_cyl

```


What happens if you reverse the order of the grouping variables?



Create a pivot (crosstab) table with class by number of cylinders:


```{r}

# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot


```

Now, unpivot it back to the original structure:

```{r}
# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

What do you notice about the structure of the unpivoted table?



## Your Turn 3.2

Making some summary tables with the {gapminder} data.

Create a table where:

* object name `lifeExp_by_continent` that has 
* the median life expectency for each continent for every year in the dataframe, 
* where the continent is the column and the year is the row.

(This is a restructuring of the table from 2.4)

```{r}


```

## Your Turn 3.3

Create a table that shows:

* GDP per capita for
* the countries Canada, United States, and Mexico are the columns, and 
* the years after 1980 are the rows

```{r}



```




***

# Take aways

* Restructure data tabulations with `pivot_wider()` and `pivot_longer`  
