---
title: "Classic Rock 25 - sales data"
output: html_notebook
---

This script uses data from Walt Hickey's article [Why Classic Rock Isn’t What It Used To Be](https://fivethirtyeight.com/features/why-classic-rock-isnt-what-it-used-to-be/) (FiveThirtyEight, 2014-07-07)



```{r setup}
library(tidyverse)
library(lubridate)

library(glue)
library()
```


## Read data

```{r}
albumlist <- read_csv("albumlist.csv")
albumlist50 <- albumlist %>%
  filter(Number < 51)
albumlist50
# fix genre (note: these categories are accurate, not precise)
genre_rev <- c("classic rock", "classic rock", "classic rock", "classic rock", "classic rock", 
               "soul-r&b-funk", "classic rock", "punk-new wave", "classic rock", "classic rock", 
               "rock & roll", "jazz", "classic rock", "classic rock", "classic rock",
               "classic rock", "alternative", "rock", "classic rock", "soul-r&b-funk",
               "rock & roll", "blues", "classic rock", "soul-r&b-funk", "soul-r&b-funk",
               "rock", "alternative", "classic rock", "classic rock", "folk rock",
               "folk rock", "classic rock", "punk-new wave", "classic rock", "rock", 
               "folk rock", "rock", "blues", "rock & roll", "classic rock", 
               "punk-new wave", "classic rock", "rock", "punk-new wave", "classic rock",
               "reggae", "jazz", "rap", "rock", "rock & roll")
albumlist50 <- cbind(genre_rev, albumlist50)
albumlist50 <- albumlist50 %>%
  mutate(price =
           case_when(
             genre_rev == "classic rock" ~ 5.99,
             genre_rev == "rock" ~ 6.49,
             genre_rev == "jazz" ~ 6.99,
             genre_rev == "blues" ~ 6.99,
             TRUE ~ 7.99
           ))
albumlist50 %>%
  group_by(genre_rev) %>%
  tally()
```

## _add price, accurate genre_


## Generate customer (store) ID table

Note: store list is defunct record retailers, active in the 80s and 90s, from [Wikipedia List of defunct Canadian companies](https://en.wikipedia.org/wiki/List_of_defunct_Canadian_companies#Consumer_retail,_including_grocery)

* HMV was left off, as it was an international brand that was late to the market

```{r}
chainlist <- c("A&A", "A&B", "Kelly's", "Sam The Record Man")
regionlist <- c("north", "east", "south", "west")
ndf = 10
storelist <- data.frame(
  store.id = 100:(ndf+99),
  chain = sample(chainlist, ndf, replace = TRUE),
  region = sample(regionlist, ndf, replace = TRUE, prob = c(0.4, 0.3, 0.2, 0.1))
  )
storelist <-
storelist %>%
  mutate(store.id = as.character(glue('X{store.id}')))
storelist
```

## Create order

### order date

https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html

```{r}
jan01 <- ymd("2019-01-01")
startdate <- as_date(floor_date(jan01, "year"))
enddate <- as_date(ceiling_date(jan01, "year") - days(1))
daterange <- tibble(orderdate = as_date(c(startdate:enddate)))

orderdate_fun <- function(){ 
daterange %>%
  sample_n(1) %>%
  pull(orderdate) 
}  

#orderdate <- orderdate_fun()
#
#orderdate
##order_final %>%
##  mutate(orderdate = orderdate)
## assign order volume based on quarter (derived from month)
#order_qtr <- quarter(orderdate)
#order_qtr
#order_mean <- case_when(
#  order_qtr == 1 ~ 4,
#  order_qtr == 4 ~ 6,
#  TRUE ~ 5
#)
#
#order_mean
order_mean_fun <- function(orderdate){
  order_qtr <- quarter(orderdate)
  order_mean <- case_when(
    order_qtr == 1 ~ 4,
    order_qtr == 4 ~ 6,
    TRUE ~ 5
  )
}
```


```{r}
store_fun <- function(){ 
storelist %>%
  sample_n(1) %>%
  pull(store.id) 
}  

#store.id <- store_fun()

```



### album_order_fun

```{r}
album_order_fun <- function(...){
  sample_n(albumlist50, 1) %>%
    select(Artist, Album, price) %>%
  mutate(format = sample(c("lp", "cass"), 1, prob = c(0.33, 0.67)),
         qty = rpois(1, 4)
  )
#         ,
#         orderdate = `orderdate`, 
#         store.id = `store.id`)
}
# TEST
#album_order_fun()
```


### multiples

```{r}
# TEST
n = rpois(1, 7)
order_final <- data.frame(Artist = as.character(),
                          Album = as.character(),
                          price = as.double(),
                          format = as.character(),
                          qty = as.integer(),
#                          orderdate = as_date(as.character()),
#                          store.id = as.character(),
                          stringsAsFactors = FALSE)  
#ls.str(order_final)
#rbind(order_final, order_fun())
#
for(i in 1:n){
  order_final[i, ] <- album_order_fun()
}
order_final
```

test orderdate and order_mean 

```{r}
orderdate <- orderdate_fun()
orderdate
order_mean <- order_mean_fun(orderdate)
order_mean
```

```{r}
order_fun_1 <- function(){
  n = rpois(1, 5)
  order_final <- data.frame(Artist = as.character(),
                          Album = as.character(),
                          price = as.double(),
                          format = as.character(),
                          qty = as.integer(),
#                          orderdate = as_date(as.character()),
#                          store.id = as.character(),
                          stringsAsFactors = FALSE)  
  
  for(i in 1:n){
    order_final[i, ] <- album_order_fun()
  }
  order_final
  }


#TEST 
order_fun_1()
```




```{r}
# now with order date and store id appended after the run

order_fun <- function(){
  
  orderdate <- orderdate_fun()

  order_mean <- order_mean_fun(orderdate)
  n = rpois(1, order_mean)
  
  store.id <- store_fun()
  
  order_final <- data.frame(Artist = as.character(),
                          Album = as.character(),
                          price = as.double(),
                          format = as.character(),
                          qty = as.integer(),
#                          orderdate = as_date(as.character()),
#                          store.id = as.character(),
                          stringsAsFactors = FALSE)  
  
  for(i in 1:n){
    order_final[i, ] <- album_order_fun()
  }
  order_final <- order_final %>%
    mutate(orderdate = orderdate,
           store.id = store.id)
  
  order_final
  }
```


```{r}
order_fun()
```





# next!

* nest this inside a tibble with the store.id and date fields

https://tidyr.tidyverse.org/reference/nest.html

https://rstudio-education.github.io/tidyverse-cookbook/transform-tables.html 





* next that inside a bigger object



