---
title: "Rolling Stone 500 - sales data"
output: html_notebook
---

This script uses the [_Rolling Stone's_ 500 Greatest Albums of All Time](https://en.wikipedia.org/wiki/Rolling_Stone%27s_500_Greatest_Albums_of_All_Time) (2012 version) to generate a fake sales order.

Data source is 

* https://data.world/notgibs/rolling-stones-top-500-albums

* download link https://query.data.world/s/sc4hbemuc4htolfcah7fs7l2oukegq


```{r setup}

library(tidyverse)

library(glue)
library(lubridate)

```


## Read data

```{r}

albumlist <- read_csv("albumlist.csv")

albumlist50 <- albumlist %>%
  filter(Number < 51)

albumlist50

# fix genre

genrelist <- c("classic rock", "classic rock", "classic rock", "classic rock", "classic rock", 
               "soul-r&b-funk", "classic rock", "punk-new wave", "classic rock", "classic rock", 
               "rock & roll", "jazz", "classic rock", "classic rock", "classic rock",
               "classic rock", "alternative", "rock", "classic rock", "soul-r&b-funk",
               "rock & roll", "blues", "classic rock", "soul-r&b-funk", "soul-r&b-funk",
               "rock", "alternative", "classic rock", "classic rock", "folk rock",
               "folk rock", "classic rock", "punk-new wave", "classic rock", "rock", 
               "folk rock", "rock", "blues", "rock & roll", "classic rock", 
               "punk-new wave", "classic rock", "rock", "punk-new wave", "classic rock",
               "reggae", "jazz", "rap", "rock", "rock & roll")

albumlist50_1 <- cbind(genrelist, albumlist50)

albumlist50_1 %>%
  mutate(price =
           case_when(
             genrelist == "classic rock" ~ 5.99,
             genrelist == "rock" ~ 6.49,
             genrelist == "jazz" ~ 6.99,
             genrelist == "blues" ~ 6.99,
             TRUE ~ 7.99
           ))

```

## _add price, accurate genre_


## Generate customer (store) ID table

Note: store list is defunct record retailers from [Wikipedia List of defunct Canadian companies](https://en.wikipedia.org/wiki/List_of_defunct_Canadian_companies#Consumer_retail,_including_grocery)


```{r}

chainlist <- c("A&A", "A&B", "HMV", "Kelly's", "Sam The Record Man")
regionlist <- c("north", "east", "south", "west")

ndf = 10

storelist <- data.frame(
  store.id = 100:(ndf+99),
  chain = sample(chainlist, ndf, replace = TRUE),
  region = sample(regionlist, ndf, replace = TRUE, prob = c(0.4, 0.3, 0.2, 0.1))
  )

storelist <-
storelist %>%
  mutate(store.id = as.character(glue('X{store.id}')))

storelist

```

## Create order

### album_order_fun

```{r}

album_order_fun <- function(...){
  sample_n(albumlist50, 1) %>%
    select(Artist, Album) %>%
  mutate(format = sample(c("lp", "cass"), 1, prob = c(0.33, 0.67)),
         qty = rpois(1, 4))
}


# TEST
album_order_fun()


```


### multiples

```{r}
# TEST

n = rpois(1, 7)
order_final <- data.frame(Artist = as.character(),
                          Album = as.character(),
                          format = as.character(),
                          qty = as.integer(),
                          stringsAsFactors = FALSE)  
#ls.str(order_final)
#rbind(order_final, order_fun())
#
for(i in 1:n){
  order_final[i, ] <- album_order_fun()
}

order_final

```



```{r}

order_fun <- function(){
  n = rpois(1, 7)
  order_final <- data.frame(Artist = as.character(),
                          Album = as.character(),
                          format = as.character(),
                          qty = as.integer(),
                          stringsAsFactors = FALSE)  
  
  for(i in 1:n){
    order_final[i, ] <- album_order_fun()
  }
  order_final
  }


```


```{r}

order_fun()

```


# next!

* nest this inside a tibble with the store.id and date fields

https://tidyr.tidyverse.org/reference/nest.html

https://rstudio-education.github.io/tidyverse-cookbook/transform-tables.html 

```{r}

store_1 <- function(){ 
storelist %>%
  sample_n(1) %>%
  pull(store.id) 
}  

order_final %>%
  mutate(store.id = store_1())


```


## lubridate

https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html

```{r}

jan01 <- ymd("2019-01-01")

startdate <- as_date(floor_date(jan01, "year"))
enddate <- as_date(ceiling_date(jan01, "year") - days(1))

daterange <- tibble(orderdate = as_date(c(startdate:enddate)))

orderdate <- function(){ 
daterange %>%
  sample_n(1) %>%
  pull(orderdate) 
}  

orderdate()

order_final %>%
  mutate(store.id = store_1(),
         orderdate = orderdate())



```



* next that inside a bigger object





