---
title: "statcan_14100383-2"
author: "Martin Monkman"
date: "05/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(cansim)
```

## data source

https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1410038302

Labour force characteristics, seasonally adjusted and unadjusted, last 5 months1, 2, 3
Frequency: Monthly

Table: 14-10-0383-02

Release date: 2022-02-04

Geography: Census metropolitan area

### license 

Statistics Canada Open License
https://www.statcan.gc.ca/en/reference/licence


## data download

use cansim package

```{r}

get_cansim_table_overview("14-10-0383-02")

tbl_1410038302 <- get_cansim("14-10-0383-02")
tbl_1410038302

```


Next chunk: create summary table with Vancouver unemployment (seasonally adjusted) with both the estimate of the number of unemployed people (in thousands—see the variable "SCALAR_FACTOR").


```{r}

# Vancouver CMA unemployment - estimate and rate, seasonally adjusted

unemployment_vancouver <- tbl_1410038302 %>% 
  filter(VECTOR == "v1235049783" | 
        VECTOR == "v1235049788")


unemployment_vancouver %>% 
  select(REF_DATE, GEO, UOM, SCALAR_FACTOR, VALUE, `Labour force characteristics`, `Data type`)

write_csv(unemployment_vancouver, "unemployment_vancouver.csv")

```

Because the "UOM" (Unit Of Measure) are Persons and Percentage, this is untidy—we have two variables in a single column, and two rows for a single observation.



```{r}

unemployment_vancouver %>% 
  select(REF_DATE, VALUE, `Labour force characteristics`) %>% 
  pivot_wider(names_from = "Labour force characteristics",
              values_from = "VALUE")


```

