---
title: "Anscombe's Quartet"
output:
  html_document:
    df_print: paged
---

Using Anscobme's Quartet to introduce principles of data visualization.

```{r setup}

library(tidyverse)

```



The quartet is included in base R, but in an untidy format:

```{r}

anscombe

```


Reshape into tidy form:

https://rpubs.com/dgrtwo/tidy-anscombe

```{r evaluate = FALSE}
# old-school `gather()` and `spread()`
anscombe_tidy <- anscombe %>%
  mutate(observation = seq_len(n())) %>%
  gather(key, value, -observation) %>%
  separate(key, c("variable", "set"), 1, convert = TRUE) %>%
  mutate(set = c("I", "II", "III", "IV")[set]) %>%
  spread(variable, value) %>%
  arrange(set)

head(anscombe_tidy)

```



## 1. Summary statistics

Each of the four sets has the same summary statistics. Let's calculate them now:


```{r}
# example
anscombe_tidy %>%
  group_by(set) %>%
  summarize(mean.x = mean(x))

```

### Your turn 1.1

Using the following functions, calculate the summary statistics of _x_ and _y_, and the correlation coefficient between _x_ and _y_, for all four of the sets in the quartet:

 | statistic | function |
 | ----       | ----      |
 | mean      | mean()   |
 | standard deviation | sd() |
 | correlation coefficient | cor() |
 
 

```{r}
# solution
anscombe_tidy %>% 
  group_by(set) %>% 
  summarize(mean(x), 
            sd(x), 
            mean(y), 
            sd(y), 
            cor(x, y))


```


## 2. Visualizing the quartet

Using the R visualization package {ggplot2}.

The template of a `ggplot()` function call looks like this:

`ggplot(data = <DATA>) + ` <BR>
`  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`  


With this example, the dataframe we're using is `anscombe_tidy` and we are plotting `x` and `y` as points.

```{r}
# example
ggplot(anscombe_tidy) + 
  geom_point(aes(x = x, y = y)) 


```


### Your turn 2.1



With `colour` to differentiate each set

```{r}
# example
ggplot(anscombe_tidy) + 
  geom_point(aes(x = x, y = y, colour = set)) 



```


## Your turn 1.3

Filter by set

```{r}
# solution
anscombe_tidy %>%
  filter(set == "I") %>%
ggplot() + 
  geom_point(aes(x, y), size = 3, colour = "purple")

```




## Your turn 1.4

Facet plot

```{r}

ggplot(anscombe_tidy) + 
  geom_point(aes(x, y)) + 
  facet_wrap(~set)

```



## Your turn 1.5

Add a trend line using the `geom_smooth` function.

* the `method = lm` indicates a "linear model", i.e. a standard regression line. {ggplot2} provides access to other smoothing algorithms.

* the `se = FALSE` turns off the "standard error" (a measure of uncertainty in the data)

```{r}

ggplot(anscombe_tidy) + 
  geom_point(aes(x, y)) + 
  geom_smooth(aes(x, y), method = lm) +
  facet_wrap(~set)

```


But that duplicates the `aes(x, y)` text ... so we can move that into the `ggplot()` function. That way, the aesthetics apply to each of the `geom_` calls.


```{r}

ggplot(anscombe_tidy, aes(x, y)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  facet_wrap(~set)

```


### References

https://rstudio-pubs-static.s3.amazonaws.com/52381_36ec82827e4b476fb968d9143aec7c4f.html 

https://rstudio-pubs-static.s3.amazonaws.com/355492_5bbde2dd99134f47b03d834b030d509b.html

https://plotlyblog.tumblr.com/post/68951620673/why-graph-anscombes-quartet 

https://rpubs.com/dgrtwo/tidy-anscombe

https://rpubs.com/turnersd/anscombe

https://www.r-bloggers.com/the-visual-difference-%E2%80%93-r-and-anscombe%E2%80%99s-quartet/


#### Excel references

https://projects.ncsu.edu/labwrite/res/gt/gt-reg-home.html


