---
title: "Plotting in R"
subtitle: "hands-on examples, with answers"
output: html_document
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. -->

```{r setup}
library(tidyverse)

library(gapminder)
```

# 1. Plotting

## Objectives

* Produce scatter plots, histograms, and time series plots using {ggplot2}.

* Set universal plot settings.
 
* Describe what faceting is and apply faceting in a `ggplot()` call.
 
* Modify the aesthetics of an existing plot (including color).

 
## REFERENCES

* _R4DS_, ["Data visualisation"](https://r4ds.had.co.nz/data-visualisation.html)

* _R4DS_, ["Exploratory Data Analysis"](https://r4ds.had.co.nz/exploratory-data-analysis.html)




## 1. Scatter plot: GDP by life expectancy

For this and some of the other examples, we will be using just the 2007 data.


### Your Turn 1.1

Create a dataframe object with the name "gapminder_2007" by filtering the gapminder data.

```{r}
# easy peasy


```


### Your Turn 1.2


Scatter plots are sometimes known as "X-Y plots" because they compare two values for the same observation. 

Create a simple scatterplot using the {gapminder} data, showing all the countries in the world for 2007, the per capita GDP on the _x_ axis, and life expectancy on the _y_. This uses the `geom_point()` function.

 
```{r}
# example: a blank plot

ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) 


# solution



# change the colour of the points


```

> b. Create a variant of the above plot, with each continent coloured differently. Try it again with shape instead of colour.

 
```{r}
# solution


# with shape instead of colour

```






***

## 2. Histogram: life expectancy

Histograms are ways to "bin" continuous variables into categories, and then count the number of observations in each bin.

The geom is `geom_histogram()`.

Like bar plots, you don't need a `y =` variable; the Y axis is the count of the number of observations.

### Your turn 2.1

Make a histogram using `gapminder_2007`, showing the count of countries by life expectancy.

```{r}
# solution


```


### Your turn 2.2

An important option in the `geom_histogram()` function is `binwidth = `. This sets the range that each bin covers. Experiment with adding bin widths of 3, 5, and 10 to the histogram

```{r}
# solution


```

Some more formatting...like you did earlier with the points, change the colour of the bars to red.

```{r}
# solution


```

"colour" is the border of the bar! Use "fill" for the inside -- make a white bar with a black border

```{r}
# solution


```


### Your turn 2.3

What do you think happens when you put a "fill" option in the `aes()` function?

```{r}
# solution


```



***

## 3. Line graph: change in life expectancy over time

Line graphs join the dots and are useful for showing change over time.

The geom is `geom_line()`.


### Your turn 3.1

Make a line graph using `gapminder`, showing the change of life expectancy for Canada between 1952 and 2007. 

Here's an example that uses the `geom_point` from earlier...change this to a line.


**plot twist**

(See what I did there?)

You can add a `ggplot()` call as part of a pipe! This means that you can filter, mutate, group, etc and then pipe those results into a plot.

Note that the data is already identified, so you don't include it after `ggplot`.

And as we saw earlier, the `aes()` could also be included in the `geom_()` function.


```{r}
# example: a point graph
gapminder %>%
  filter(country == "Canada") %>%
  # now the ggplot 
  ggplot(aes(x = year, y = lifeExp)) +  # IMPORTANT - it's a "+", not a "%>%"
  geom_point()

# solution 


```


### Your turn  3.2

Modify the code you wrote in #3.1 to show the life expectancy for Canada, Oman, and Japan. You will still have year and life expectancy as the point value represented on the axes -- how will you separate the countries?

(In 2007, Japan had the longest life expectancy of any country, and in the years between 1952 and 2007, Oman had the biggest increase--a change of 38 years! The code for how I determined this is at the bottom of the page.) 

```{r}

country_list <- c("Canada", "Oman", "Japan")



```

### Your turn  3.3

Layering geoms ... it is possible to add geoms to a plot. 

See what happens when you use the code from 3.2, with `geom_line`, and then add the `geom_point` from the example in 3.1.

```{r}

# solution

```


### BONUS CALCULATIONS


Biggest change in life expectancy

```{r}

gapminder %>%
  filter(year %in% c(1952, 2007)) %>%
  group_by(country) %>%
  mutate(lifeExp_change = lifeExp - lag(lifeExp)) %>%
  arrange(desc(lifeExp_change))


```




### REFERENCE

Some of these examples are modified from http://euclid.psych.yorku.ca/www/psy6135/tutorials/gapminder.html. Thanks Michael!











