---
title: "Plotting categorical variables"
subtitle: "hands-on examples"
output: html_document
---

<!-- This file by Martin Monkman is licensed under a Creative Commons Attribution 4.0 International License. -->

```{r setup}
library(tidyverse)

library(gapminder)
```

# 1. Plotting categorical variables

## Objectives

* Produce bar charts and other categorical plots using {ggplot2}.

* Set universal plot settings.
 
* Extend learning of {ggplot2} functions

## REFERENCES

* https://www.r-graph-gallery.com/267-reorder-a-variable-in-ggplot2.html

* https://stat545.com/factors-boss.html#reorder-factors


## Set up the data



Create a dataframe object with the name "gapminder_2007" by filtering the gapminder data.

```{r}

gapminder_2007 <- filter(gapminder, year == 2007)

```



## 1. Bar plot: countries in each continent


Bar plots are often used to visualize the distribution of a discrete variable. In this case, we will show how many countries there are in each continent.

* the **geom** is `geom_bar`

* map the `x` variable to `continent`

* there is no `y` variable! {ggplot} will count the number of observations in each category

### Your Turn 1.1


```{r}
# solution
ggplot(gapminder_2007, aes(x = continent)) +
  geom_bar()
```

### Your Turn 1.2

Note that `colour` (or `color`) won't work on a bar! That's for points and lines.

For something that occupies a block of space--such as a bar or pie chart--you need to use `fill`.

Add the `fill` attribute to `continent` to the code you wrote above. (Yes, you'll be specifing continent twice!)

```{r}
# solution
ggplot(gapminder_2007, aes(x = continent, fill = `continent`)) +
  geom_bar()


```


This is the default palette. It might be a bit too vibrant for your eyes...don't worry, we will learn to fix that later.





### Your turn 1.3

Another {ggplot2} feature is that every plot is an object. If you want to save a basic version of your plot and continue to tinker with it, you can assign that basic version to an object name, and just add to it.

It would look something like:

`mybar <- ggplot() + geom_()`

Followed by

`mybar + `



```{r}
# solution


mybar <- ggplot(gapminder_2007, aes(x = continent, fill = `continent`)) +
  geom_bar()


```



### Your turn 1.4 



It's possible to turn this on its side, so that the country labels are on the left, and the bars run left-to-right instead of up-and-down.

To do this, add the `coord_flip()` function to the chart object you assigned above (you might have called it `mybar`).




```{r}
# solution

mybar + coord_flip()

```


## 2. Sorting factors in a plot

To sort the factors in a plot, we first mutate the variable that contains the factor variable that will plot.

In this example, we use the same code that we saw in hands-on 3.3, Your Turn 3.2. Other sort functions (such as `fct_relevel`) could also be used. 

```{r}
gapminder %>%
  # sort the factors
  mutate(continent = continent %>% 
           fct_infreq() %>% 
           fct_rev()) %>% 
  # then plot
  ggplot(aes(x = continent)) +
    geom_bar() +
  coord_flip()

```


### Your turn 2.1


In this example, we will plot the four countries in our BIDA group by life expectancy.

```{r}
h_countries <- c("Belgium", "India", "Denmark", "Albania")  # see what I did there?

h_gap <- gapminder %>%
  filter(country %in% h_countries) %>%
  mutate(country = country %>% 
  fct_relevel("Belgium", "India", "Denmark", "Albania"))  

h_gap %>%
  filter(year == 2007) %>%
  ggplot(mapping = aes(x = country, y = lifeExp)) +
  geom_point(size = 6)

```


Now it's your turn: sort by life expectancy:

```{r}

h_gap %>%
  filter(year == 2007) %>%
  mutate(country = country %>% 
           fct_reorder(lifeExp) %>% 
           fct_rev()) %>% 
  # now plot
  ggplot(mapping = aes(x = country, y = lifeExp)) +
  geom_point(size = 6)


```






## REFERENCE

Some of these examples are lifted directly from http://euclid.psych.yorku.ca/www/psy6135/tutorials/gapminder.html


See also

* https://www.r-graph-gallery.com/267-reorder-a-variable-in-ggplot2.html

* https://stat545.com/factors-boss.html#reorder-factors








