---
title: "More plot strategies"
author: "Martin Monkman"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r, setup, message=FALSE}
# tidyverse
library(tidyverse)
library(lubridate)
#
library(scales)     # extending {ggplot2}
#
library(gapminder)
```



## Small multiples

One important strategy in plots is to "eliminate clutter".

Here's an example of a cluttered plot:

```{r}

gapminder %>% 
  filter(continent == "Americas") %>% 
  #
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line()

```

A strategy for reducing the clutter is to split this into sub-plots. The way to do this is to use _facets_, sometimes refered to as _small multiples_.

### `facet_wrap()` 

In this case, we replacing the `colour = ` argument in the `aes()` with a separate `facet_wrap()` function:

```{r}

gapminder %>% 
  filter(continent == "Americas") %>% 
  #
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap( ~ country)



```


Another example, using {gapminder} GDP per capita.

Note that even though the structure of the syntax is a bit different, we end up with a plot.

```{r}
p <- ggplot(gapminder, aes(x = year, y = gdpPercap))

p + geom_line(aes(group = country))
            
            
```


Let's take that plot object `p` and facet by continent:
 
```{r}

p + geom_line(aes(group = country)) +
  facet_wrap(~continent)

```

That's an improvement, but it can be still better. What we want to really do is compare each continent's GDP--it would be better to have a single row. To do that we add the number of columns we want:

```{r}
p + geom_line(aes(group = country)) +
  facet_wrap(~continent, ncol = 5)

```

### `facet_grid()` 

This function allows you to arrange your facets by two dimensions.

For this, we'll use the {mpg} data.

```{r}

mpg

```

```{r}
p <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()

p 
```

Let's grid by a single variable, `cyl`. Note that the dot `.` is the row specification, so `cyl` will be columns.

```{r}
p + facet_grid(. ~ cyl)
```

And now `drv` as the rows:

```{r}
p + facet_grid(drv ~ .) 
```

And finally, a facet grid by both `drv` and `cyl`:

```{r}
p + facet_grid(drv ~ cyl) 
```


## Using colour

Colour can be very effective in bringing attention to a particular group.

In this example, we will colour the positive values differently than the negative.

For the data, we'll use a data set showing life expectancy in the United States and the other OECD countries, the difference between them, and the variable `hi_lo` to show whether the OECD average is above or below the USA.

```{r}
socviz::oecd_sum
```

Our basic plot:

```{r}
p <- ggplot(data = socviz::oecd_sum,
            mapping = aes(x = year, y = diff)) +
        geom_col()

p

```

Because this is a column chart, we can add the `fill = ` argument in the `aes()` to colour the columns based on the values in a variable. 

```{r}
p <- ggplot(data = socviz::oecd_sum,
            mapping = aes(x = year, y = diff, fill = hi_lo)) +
        geom_col()

p

```


And now with more formatting. Note that removing the legend is a way to reduce the amount of clutter.

```{r}

p + 
  # remove the legend
  guides(fill = FALSE) +
  # titles
  labs(x = NULL, y = "Difference in Years",
       title = "The US Life Expectancy Gap",
       subtitle = "Difference between US and OECD
                   average life expectancies, 1960-2015",
       caption = "Data: OECD. After a chart by Christopher Ingraham,
                  Washington Post, December 27th 2017.")
```

Souce: [Kieran Healy, _Data Visualization: A Practical Introduction_](http://socviz.co/groupfacettx.html#avoid-transformations-when-necessary)

-30-