---
title: "Getting R & RStudio up and running on your own computer"
author: "BIDA 302"
#date: "2019/12/07"
output:
  xaringan::moon_reader:
#    css: ["default", "css_files/eigengrau.css", "css_files/eigengrau-fonts.css"]    
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


<!-- 
This file by Martin Monkman 
is licensed under a Creative Commons Attribution 4.0 International License
https://creativecommons.org/licenses/by/4.0/  
-->




```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#16161D",
  secondary_color = "#6e6e91",
  inverse_header_color = "#e2e2e9",
  header_font_google = google_font("Raleway"),
  text_font_google   = google_font("Raleway"),
  code_font_google   = google_font("Fira Mono")
)
```


## Three components

1. R

2. RStudio

3. packages


---

## How-to

_It depends_

* operating system (Windows, Mac, or Linux)

* administrative privilege & network permissions

---

## Best source:

[_R Cookbook, 2nd edition_, "Getting Started and Getting Help"](https://rc2e.com/gettingstarted)

* [1.1 Downloading and Installing R](https://rc2e.com/gettingstarted#recipe-id001)

* [1.2 Installing RStudio](https://rc2e.com/gettingstarted#recipe-id002b)

* [3.10 Installing Packages from CRAN](https://rc2e.com/navigatingthesoftware#recipe-id012)


---

## Packages: {tidyverse} 

[tidyverse: Easily Install and Load the 'Tidyverse'](https://cran.r-project.org/web/packages/tidyverse/index.html)

* many of the packages we've used, including 

  - wrangling with {dplyr} & {tidyr}

  - visualization with {ggplot2}

  - importing with {readr} & {readxl}

  - working with variables with {forcats}, {lubridate} & {stringr}

```
install.packages("tidyverse")
```

---

## Packages: others 

[janitor: Simple Tools for Examining and Cleaning Dirty Data](https://cran.r-project.org/web/packages/janitor/index.html)

[glue: Interpreted String Literals](https://cran.r-project.org/web/packages/glue/index.html)





