---
title: 'BIDA302 - assignment #1: 4 & 7'
subtitle: 'solutions'
output: html_document
---


Package load:
```{r setup}
library(tidyverse)
# data files
library(nycflights13)

```




***

### 4. Religion and income data (USA)
> (2 marks)

> The table `relig_income` is in {tidyr}.
>
>a. Load this table, and take a look at it. What tidy principle(s) does this table violate? (See the ["Tidy Data" chapter](https://r4ds.had.co.nz/tidy-data.html) in _R4DS_., and ["Tidy Data for Efficiency, Reproducibility, and Collaboration"](https://www.openscapes.org/blog/2020/10/12/tidy-data/))
>

```{r}
tidyr::relig_income
```

It violates Principle #1, each variable must have its own column. In this example, the "income" variable is spread across 10 columns.

>
>b. Using the functions in {tidyr}, arrange this table in such a way that it is no longer messy.


```{r}

relig_income_long <- relig_income %>%
  pivot_longer(-religion, names_to = "income_cat", values_to = "count")

relig_income_long
```

And to pivot it back to wide layout:


```{r}

relig_income_long %>%
  pivot_wider(names_from = "income_cat", values_from = "count")

```


***

### 7. Car mileage - pivot table
> (2 marks)

> In Lesson 1, we used the {mpg} package to calculate the average displacement by class of car:

```{r}
mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```
> We then created a pivot (crosstab) table with class by number of cylinders, and then unpivoted that table:

```{r}
# summary table - class by cylinder
class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) %>%
  arrange(desc(displ_mean))

# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot

# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

> a. What happens when you repeat the first chunk of code to calculate the average displacement of each class of cars but with the new object `displ_class_by_cyl`?

```{r}

displ_class_by_cyl %>%
  group_by(class) %>%
  summarise(mean = mean(displ_mean))

```

You get "NA" for all but one of the class variables! 

This is because the `mean()` function requires valid values.

> b. Solve this problem so that the table looks like the first one in this question. Use any and all of the reference materials at your disposal!

The way to solve this is to use the option `na.rm = TRUE` (`rm` means "remove") in the function call.


```{r}

displ_class_by_cyl %>%
  group_by(class) %>%
  summarise(mean = mean(displ_mean, na.rm = TRUE))

```
