---
title: 'BIDA302 - assignment #1'
subtitle: 'solutions'
output: html_document
---


Package load:
```{r setup}
library(tidyverse)
# data files
library(nycflights13)

```

## YOUR NAME HERE



## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have submitted your assignment, I will be hitting the "Knit" button—please ensure that the output renders correctly.



## Questions


### 1. Assign a variable
> (1 mark)

> Why does this code chunk not work?

```{r eval=FALSE}
stringlenght <- 10

stringlength
```

> Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)

**Answer:**

The variable is spelled differently in each case.


### 2. Reducing the size of your data set
> (1 mark)

> Starting with the {mpg} data frame, use `filter()` and `select()` to create a shorter form of this table that contains the following records and variables:
>
>* compact and subcompact Volkswagens from the model year 1999
>
>* manufacturer, model, mileage in both city and highway 

```{r}
# one possible solution
mpg %>%
  filter(manufacturer == "volkswagen",
         class %in% c("compact", "subcompact"),
         year == 1999) %>%
  select(manufacturer, model, cty, hwy)

```

***

### 3. Selecting variables
> (1 mark)

> What happens if you include the name of a variable twice in a `select()` call? Demonstrate using the {gapminder} data.


The variable is _not_ duplicated.

```{r}
library(gapminder)

gapminder %>%
  select(country, year, lifeExp, year)
  

```



***

### 4. Religion and income data (USA)
> (2 marks)

> The table `relig_income` is in {tidyr}.
>
>a. Load this table, and take a look at it. What tidy principle(s) does this table violate? (See the ["Tidy Data" chapter](https://r4ds.had.co.nz/tidy-data.html) in _R4DS_., and ["Tidy Data for Efficiency, Reproducibility, and Collaboration"](https://www.openscapes.org/blog/2020/10/12/tidy-data/))
>

```{r}
tidyr::relig_income
```

It violates Principle #1, each variable must have its own column. In this example, the "income" variable is spread across 10 columns.

>
>b. Using the functions in {tidyr}, arrange this table in such a way that it is no longer messy.


```{r}

relig_income_long <- relig_income %>%
  pivot_longer(-religion, names_to = "income_cat", values_to = "count")

relig_income_long
```

And to pivot it back to wide layout:


```{r}

relig_income_long %>%
  pivot_wider(names_from = "income_cat", values_from = "count")

```

***


### 5. R Markdown
> (1 mark)

> The following chunk of code produces a chart. When this notebook is previewed, the R code inside the chunk is also rendered.
>
> Enter a chunk option so that the chart shows in the output, but the code does not.
>
> (Hint: use the RMarkdown cheatsheet to find the solution!)

```{r}
# hide this code!
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```

The solution is to put the option "echo = FALSE" after the "r" in the opening bit of the code.


```{r, echo=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

```





***

### 6. Flight data
> (4 marks)

> For Q3a-3d, use the data `nycflights13::flights`
>
> Pay close attention to how the variables are coded.
>
> Find all flights that:
>
>a. Had an arrival delay of two or more hours

Note that variable `arr_delay` is measured in minutes, so we need to find the flights where that's greater than or equal to 120 minutes.

```{r}
filter(flights, arr_delay >= 120)
```


>b. Flew to Houston (there are two airports, with the codes IAH or HOU)

```{r}

filter(flights, dest == "IAH" | dest == "HOU")

```

Note that you could also use `%in%`

>c. Were operated by United, American, or Delta

This one takes a bit of effort--you have to find the codes that match the values in the variable `carrier` with the airline names.

The {nycflights13} package has a number of tables, including `airlines`

```{r}
nycflights13::airlines
```

From this, we can find:

* United == UA

* American == AA

* Delta == DL

```{r}

filter(flights, carrier %in% c("AA", "DL", "UA"))

```


>d. Departed in summer (July, August, and September)

The variable `month` is numeric, so we have to use greater than and less than to define the range.

```{r}

filter(flights, month >= 7, month <= 9)

```


Using the `:` operator would allow us to make a range. 

```{r}

filter(flights, month %in% 7:9)

```

You could also use "or" `|`, or `between()`


```{r}

filter(flights, month == 7 |
         month == 8 |
         month == 9)

filter(flights, between(month, 7, 9))

```

***

### 7. Car mileage - pivot table
> (2 marks)

> In Lesson 1, we used the {mpg} package to calculate the average displacement by class of car:

```{r}
mpg %>%                                
  group_by(class) %>%                   
  summarise(displ_mean = mean(displ)) 

```
> We then created a pivot (crosstab) table with class by number of cylinders, and then unpivoted that table:

```{r}
# summary table - class by cylinder
class_by_cyl <- mpg %>%
  group_by(class, cyl) %>%
  summarise(displ_mean = mean(displ)) %>%
  arrange(desc(displ_mean))

# pivot table (wide)

class_by_cyl_pivot <- class_by_cyl %>%
  pivot_wider(names_from = cyl, values_from = displ_mean)

class_by_cyl_pivot

# and back to longer...

displ_class_by_cyl <- class_by_cyl_pivot %>%
  pivot_longer(-class, names_to = "cyl", values_to = "displ_mean")

displ_class_by_cyl

```

> a. What happens when you repeat the first chunk of code to calculate the average displacement of each class of cars but with the new object `displ_class_by_cyl`?

```{r}

displ_class_by_cyl %>%
  group_by(class) %>%
  summarise(mean = mean(displ_mean))

```

You get "NA" for all but one of the class variables! 

This is because the `mean()` function requires valid values.

> b. Solve this problem so that the table looks like the first one in this question. Use any and all of the reference materials at your disposal!

The way to solve this is to use the option `na.rm = TRUE` (`rm` means "remove") in the function call.


```{r}

displ_class_by_cyl %>%
  group_by(class) %>%
  summarise(mean = mean(displ_mean, na.rm = TRUE))

```
