---
title: "BIDA302 - assignment #4"
output: html_document
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> All of the packages and data you will need for this assignment are in the rstudio.cloud project. There is no need for you to download and install anything.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


## 1. Joins with HR data
> 
> The Lahman baseball database contains a comprehensive listing of annual summaries for the major leagues of baseball, going back to 1871. 
> 
> For this question, smaller versions of three of the files have been created:
> 
> * "Master_2000s.csv" -- all active players from the years 2000 to 2018.
> 
> * "Salaries_2016.csv" -- the annual summary of each player's compensation, for the 2016 season.
> 
> * "Batting_2016.csv" -- the annual summary of each player's batting, for the 2016 season. (Note #1: that this omits pitchers who did not come to bat in the 2016 season. Note #2: players who were traded between teams mid-season show up with a record for their performance of each team.)
> 
> Think of these as the sorts of files that you would find in any human resources department: 
>
> * the Master file contains the personal details of each individual employed by the enterprise, for the years 2000 through 2018.
>
> * Batting contains in-depth performance measures for the employees in a specific task group
>
> * Salaries is an incomplete tables showing the compensation of some of the employees


### 1a. Read in the three files
> (1 mark)

Read in the three files, saving each as a separate object.

```{r}
# solution

```

### 1b. Understanding the tables
> (2 marks)

> i. What is the key variable that links the three tables?


> ii. Which table has the primary key? What is the reason behind your answer?


> iii. Why would the designers of the table put only one variable in the Batting table that identifies a player?



### 1c. Joins
> (3 marks)

> Join the tables in such a way so that you can tabulate the number of batters that played on each team in 2016, by birth country.
>
> Note: because some of the non-key variables exist in multiple tables, {dplyr} will rename variables as `name.x` and `name.y` -- watch out for this! Take a look at your variables after you've completed the join.

```{r}
# solution


# alternate version, using a pipe





```




## 2. Data organization in spreadsheets

>(6 marks)

> Read the following article: 

> Karl W. Broman and Kara H. Woo, "Data Organization in Spreadsheets", _The American Statistician_, Vol 72, Issue 1: Special Issue on Data Science, 2018

> * it is available on the moodle site or from this link: https://www.tandfonline.com/doi/full/10.1080/00031305.2017.1375989


> 2.a  What are basic principles for using spreadsheets for good data organization?



> 2.b  What are good approaches for handling dates in spreadsheets?



---

### Bonus Marks!
> (2 marks)

> Imagine that time travel is possible...
 
> You have just been transported back four weeks to the day before the course started. What is one thing you would tell your past self about this course?






