---
title: "BIDA302 - assignment #4"
subtitle: "solutions"
output: html_document
---

## YOUR NAME HERE


Package load:
```{r setup}
library(tidyverse)
```


## Introduction

> For this homework assignment, please write your answer for each question after the question text but before the line break before the next one.

> All of the packages and data you will need for this assignment are in the rstudio.cloud project. There is no need for you to download and install anything.

> In some cases, you will have to insert R code chunks, and run them to ensure that you've got the right result.

> Use all of the R Markdown formatting you want! Bullets, bold text, etc. is welcome.

> Once you have finished your assignment, create an HTML document by "knitting" the document using either the "Preview" or "Knit" button in the top left of the script window frame


## Questions


## 1. Joins with HR data
> 
> The Lahman baseball database contains a comprehensive listing of annual summaries for the major leagues of baseball, going back to 1871. 
> 
> For this question, smaller versions of three of the files have been created:
> 
> * "Master_2000s.csv" -- all active players from the years 2000 to 2018.
> 
> * "Salaries_2016.csv" -- the annual summary of each player's compensation, for the 2016 season.
> 
> * "Batting_2016.csv" -- the annual summary of each player's batting, for the 2016 season. (Note #1: that this omits pitchers who did not come to bat in the 2016 season. Note #2: players who were traded between teams mid-season show up with a record for their performance of each team.)
> 
> Think of these as the sorts of files that you would find in any human resources department: 
>
> * the Master file contains the personal details of each individual employed by the enterprise, for the years 2000 through 2018.
>
> * Batting contains in-depth performance measures for the employees in a specific task group
>
> * Salaries is an incomplete tables showing the compensation of some of the employees


### 1a. Read in the three files
> (1 mark)

```{r}
# solution
Master_2000s <- read_csv("Master_2000s.csv")

Batting_2016 <- read_csv("Batting_2016.csv")

Salaries_2016 <- read_csv("Salaries_2016.csv")

```

### 1b. Understanding the tables
> (2 mark)

> i. What is the key variable that links the three tables?

A: playerID

The ID is unique so they don’t need any other way to identify the players. Any other table can be joined with the players table for more information about a player.
 
> ii. Which table has the primary key? What is the reason behind your answer?

A: `Master_2000s.csv` has the primary key. It lists all individual players. The other dataframes potentially include multiple entries for each player ID. For example, a player might be traded mid-season and have two records in `Batting_2016`, one for each team played for that season. 

> iii. Why would the designers of the table put only one variable in the Batting table that identifies a player?

A: Since all of the data in `Master` is constant (name, birthdate, etc) there is no need to repeat those variables in the other tables where there are multiple records for the same player. It also means that this contsistent data only has to be edited in one location, not across all of the tables.

### 1c. Joins
> (3 marks)

> Join the tables in such a way so that you can tabulate the number of batters that played on each team in 2016, by birth country.
>
> Note: because some of the non-key variables exist in multiple tables, {dplyr} will rename variables as `name.x` and `name.y` -- watch out for this! Take a look at your variables after you've completed the join.

```{r}
# solution
MLB_HR_data1 <- 
left_join(Batting_2016, Master_2000s, by = "playerID")

MLB_HR_data <- 
left_join(MLB_HR_data1, Salaries_2016, by = "playerID")


# alternate version, using a pipe
MLB_HR_data <- 
Batting_2016 %>%
  left_join(Master_2000s, by = "playerID") %>% 
  left_join(Salaries_2016, by = "playerID")

MLB_HR_data

MLB_HR_data %>% 
  group_by(teamID.x, birthCountry) %>% 
  tally() 


# NOTES
#
# * both tables have the variable "playerID"
# 
# * we want a `left_join`: all the observations from the active season, and adding the country of birth variable from "Master"
#
# * an `inner_join` will work for "Batting_2016", with added "Master_2000s" because all players are listed in both. Adding "Salaries_2016" using an inner join will drop the records of players for whom there is no salary data


```


### 1d. Filtering joins

> (3 marks)

> How many records are in `Batting_2016` and `Salaries_2016`?
>
> Earlier it was noted that the batting table _omits_ players ... but it's the longer of the two!
> 
> Use one of the filtering joins to see how many players are listed in `Batting_2016` that don't show up in `Salaries_2016`.


```{r}
# solution

anti_join(Batting_2016, Salaries_2016, key = "playerID")

anti_join(Batting_2016, Salaries_2016, key = "playerID") %>% 
  group_by(playerID) %>% 
  tally()

# this is an anti_join -- it will filter "Batting_2016" and leave only those records where there is on matching record in "Salaries_2016

```


### 1e. Plot!

(3 marks)

> Use your exploratory data analysis skills to create two visualizations to see what might explain player compensation (salary). 
> First, compare it to the age of player, and then to the player's performance, using the crude measure of runs scored (the "R" variable in the batting table).
> Write a short explanation of what you see in the charts.

```{r}
ggplot(MLB_HR_data, aes(x = birthYear, y = salary)) +
  geom_point() +
  xlim()
```



```{r}
ggplot(MLB_HR_data, aes(x = R, y = salary)) +
  geom_point()
```



---

### Bonus Marks!
> (2 marks)

> Imagine that time travel is possible...
 
> You have just been transported back four weeks to the day before the course started. What is one thing you would tell your past self about this course?



## REFERENCE

The source of the Lahman baseball data table is the R package [{Lahman}](https://cran.r-project.org/web/packages/Lahman/index.html), which makes more accessible the data from [Sean Lahman's historical baseball database](http://www.seanlahman.com/baseball-archive/statistics/). I modified the Batting and Master tables as follows:

```{r evaluate = FALSE}
# 
library(Lahman)
library(readr)

Batting_2016 <- Batting %>%
 filter(yearID == 2016)

write_csv(Batting_2016, "Batting_2016.csv")
# ---
Master_2000s <- Master %>%
 filter(finalGame > "2000-01-01")

write_csv(Master_2000s, "Master_2000s.csv")
# ---
Salaries_2016 <- Salaries %>% 
filter(yearID == 2016)

write_csv(Salaries_2016, "Salaries_2016.csv")



```







